---
title: "Correlated batch effects on class proportions: global, partial-pooling and GP weight priors"
author: "batchmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correlated batch effects on class proportions: global, partial-pooling and GP weight priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5)
```

## Purpose

`batch_weight_prior` controls how the class/cluster mixing weights are
allowed to vary across batches: `"global"` ties every batch to one shared
weight vector, `"partial_pooling"` lets each batch have its own weight with
no assumed order or structure (only exchangeable shrinkage towards a common
mean), and `"gp"` places a Gaussian process over a batch coordinate (e.g.
collection time), so batches close together are expected to have similar
proportions. The getting-started and covariance-model vignettes cover the
mean/covariance side of batch correction; this one is specifically about
the *proportions*, using a
scenario designed so the three choices disagree with each other.

**Scenario.** Twelve assay batches (e.g. an ELISA seroprevalence panel)
are collected at irregular points across a two-year surveillance window -
enough batches to trace out an actual epidemic curve, not just a single
before/after step. The true proportion of positives follows that curve:

- **Baseline**: low and quiet before anything has happened.
- **Outbreak onset**: a genuine discontinuity - prevalence jumps, it does
  not ramp up smoothly - followed by **spread**, a rapid rise to a peak.
- **Intervention**: prevalence declines from the peak.
- **Solution**: the decline settles at a controlled level - still above
  the original baseline (an outbreak does not usually erase itself back to
  zero), but well below the peak.
- **Second wave**: a second discontinuity later in the window - a smaller
  jump than the first, since some combination of immunity and awareness
  usually blunts a second wave - followed by its own partial rise and
  decline.

Two genuine discontinuities, not one, is deliberate: it gives
`partial_pooling` and `gp` two separate chances to be tested against a
regime change, rather than one, and lets the second (smaller, later) jump
be checked against the first.

This is what the three `batch_weight_prior` choices disagree about:
`"global"` cannot represent any of this curve, `"partial_pooling"` can
represent a jump immediately with no assumption about batch order, and
`"gp"` assumes nearby batches have similar proportions - an assumption
each genuine discontinuity violates right at its boundary, making the
batches immediately after each jump the hardest for `"gp"` to get right,
even though it is the only one of the three built to use real
collection-time information at all.

Separately, and **not** tied to collection time at all, each batch's own
assay readout carries independent measurement noise (an additive shift and
a multiplicative scale, standing in for reagent-lot and instrument
calibration variation) - a batch run in month 3 is exactly as likely to be
a noisy run as one run in month 20. This is a deliberate change from an
early design of this vignette, where both the assay noise and the true
prevalence varied smoothly with time together: with the two entangled, it
was never clear whether `"gp"`'s use of collection time was tracking the
epidemic curve it is meant to track, or incidentally also picking up
assay drift that happened to correlate with time. Decoupling them means
any advantage `"gp"` shows below is specifically about the proportions,
not a side effect of the assay noise also being smooth in time.

```{r libs, message=FALSE}
library(batchmix)
library(ggplot2)
```

## 1. Simulate the data

`generateBatchData()` cannot produce this scenario (it has no notion of a
time-varying batch effect or batch-specific class proportions), so the data
is simulated directly. Batch collection times are irregular - drawn at
random over a 24-unit window, not evenly spaced - since that is what
`batch_weight_prior = "gp"` is built for, and it is worth noting
that the model uses genuine inter-batch distances, not just batch order.

```{r simSetup}
set.seed(2024)

B <- 12
time_points <- sort(round(runif(B, 0, 24), 2))
cat("Batch collection times:", time_points, "\n")
cat("Gaps between consecutive batches:", round(diff(time_points), 2), "\n")
```

```{r epiCurve}
# The epidemic curve described above, as a piecewise-linear function of
# time with two genuine discontinuities (the jumps at outbreak_t and
# wave2_t are between the two adjacent *branches* of the ifelse(), not a
# smooth transition): baseline -> outbreak jump -> spread -> intervention
# decline -> solution plateau -> second-wave jump -> rise -> partial
# decline. Every batch's true prevalence is this curve evaluated at that
# batch's own (irregular) collection time, so which batches happen to fall
# near a discontinuity is data-driven, not hand-picked.
outbreak_t <- 6 # first discontinuity: outbreak onset
peak_t <- 10 # spread ends / intervention-driven decline begins
solution_t <- 16 # decline reaches the post-intervention "solution" level
wave2_t <- 18 # second discontinuity: second-wave onset
wave2_peak_t <- 21 # second wave's own rise ends
end_t <- 24

p_baseline <- 0.05
p_outbreak_start <- 0.30 # jumps straight here at outbreak_t, no ramp
p_peak <- 0.70
p_solution <- 0.15 # above p_baseline, well below p_peak
p_wave2_start <- 0.40 # jumps straight here at wave2_t, no ramp
p_wave2_peak <- 0.55 # a smaller second wave than the first
p_final <- 0.25 # partial resolution by the end of the window

true_prevalence <- function(t) {
  ifelse(t < outbreak_t, p_baseline,
    ifelse(t < peak_t, p_outbreak_start + (p_peak - p_outbreak_start) * (t - outbreak_t) / (peak_t - outbreak_t),
      ifelse(t < solution_t, p_peak + (p_solution - p_peak) * (t - peak_t) / (solution_t - peak_t),
        ifelse(t < wave2_t, p_solution,
          ifelse(t < wave2_peak_t, p_wave2_start + (p_wave2_peak - p_wave2_start) * (t - wave2_t) / (wave2_peak_t - wave2_t),
            p_wave2_peak + (p_final - p_wave2_peak) * (t - wave2_peak_t) / (end_t - wave2_peak_t)
          )
        )
      )
    )
  )
}

true_p1 <- true_prevalence(time_points)

# The batch immediately at-or-after each discontinuity - the hardest single
# batch for "gp" to get right at that jump, and the pair this vignette's
# diagnostics and predictive checks below focus on, generalised to
# whichever batches this random draw of irregular times actually produced
# (rather than a fixed index that would only be valid for one specific
# draw).
outbreak_batch_idx <- which(time_points >= outbreak_t)[1]
wave2_batch_idx <- which(time_points >= wave2_t)[1]

cat(
  "Batch", outbreak_batch_idx, "(t =", time_points[outbreak_batch_idx],
  ") is first to land at/after the outbreak jump at t =", outbreak_t,
  "- true P(positive) goes from", round(true_prevalence(outbreak_t - 0.01), 2),
  "just before to", round(true_p1[outbreak_batch_idx], 2), "there.\n"
)
cat(
  "Batch", wave2_batch_idx, "(t =", time_points[wave2_batch_idx],
  ") is first to land at/after the second-wave jump at t =", wave2_t,
  "- true P(positive) goes from", round(true_prevalence(wave2_t - 0.01), 2),
  "just before to", round(true_p1[wave2_batch_idx], 2), "there.\n"
)
```

A GP has no time-based warning that either jump is about to happen: right
up until the batch that lands at or after each discontinuity, the
smoothest reasonable extrapolation from every earlier batch is "more of
the same".

```{r simSetupContinued}
K <- 2 # seronegative (0) / seropositive (1)
P <- 2 # two correlated assay readouts

# True, batch-free class means: fixed - the biology itself does not change
# over time, only what is measured does. Only 2 units apart per feature
# (not the much wider gap - e.g. 5 units - a first attempt might pick):
# with classes this close, an individual item's likelihood is genuinely
# ambiguous near the boundary, which is what lets the batch-specific class
# weight actually swing a classification - at a wide separation, the
# likelihood alone resolves almost every item regardless of which weight
# prior is used, and the three choices this vignette compares would look
# nearly identical downstream (a weaker, less informative demonstration).
mu_true <- rbind(c(0, 0), c(2, 2))
class_sd <- 1.0

# Per-batch assay noise (reagent-lot/instrument calibration variation):
# drawn independently per batch, deliberately NOT a function of collection
# time (see the Purpose section above for why) - an additive shift and a
# multiplicative scale on top of each batch's readout. Only the scale
# affects how separable the two classes look (an additive shift moves both
# classes by the same amount). The scale is 1 (S_loc's own lower bound)
# plus a log-normal excess: mostly modest inflation (median exp(-1) =
# 0.37, so a typical batch's variance is inflated by a bit over a third),
# but log-normal's right tail lets an occasional batch be substantially
# noisier than the rest, which a bounded distribution (e.g. abs(rnorm())
# alone) would not produce as naturally.
batch_shift_true <- rnorm(B, mean = 0.3, sd = 0.6)
batch_scale_true <- 1 + exp(rnorm(B, mean = -1.0, sd = 0.6)) # S_loc has support [1, Inf)
```

```{r simGenerate}
# Batch sizes vary too - a real surveillance programme does not draw
# exactly the same number of samples every time it runs the assay. Each
# batch's own n_per_batch[b] is what generates and indexes that batch's
# rows below; N is the SUM across batches, not a per-batch count times B
# (that would badly overcount the total, and every array below has to be
# sized by the true total or the indexing runs off the end).
n_per_batch <- rpois(B, 80)
N <- sum(n_per_batch)

X_obs <- matrix(0, N, P)
X_true <- matrix(0, N, P)
labels_true <- integer(N)
batch_vec <- integer(N)

row_i <- 1
for (b in seq_len(B)) {
  n1 <- rbinom(1, n_per_batch[b], true_p1[b])
  n0 <- n_per_batch[b] - n1
  cls <- c(rep(0, n0), rep(1, n1))
  for (k in cls) {
    z <- rnorm(P, mean = mu_true[k + 1, ], sd = class_sd)
    X_true[row_i, ] <- z
    # batch_scale_true is a variance-inflation factor, matching how the
    # fitted model applies S_b to the covariance linearly (Var = class_sd^2
    # * batch_scale) - hence sqrt() here, not a direct SD multiplier.
    X_obs[row_i, ] <- (z - mu_true[k + 1, ]) * sqrt(batch_scale_true[b]) +
      mu_true[k + 1, ] + batch_shift_true[b]
    labels_true[row_i] <- k
    batch_vec[row_i] <- b - 1
    row_i <- row_i + 1
  }
}
```

**Semi-supervised labels: the extremes, not a random sample.** Rather than
labelling a random subset (`generateBatchData()`'s usual `frac_known`),
here the "confirmed" items are the most extreme observed values within
each true class, per batch - the items a lab would realistically have
independent confirmatory testing for (e.g. a panel's positive/negative
controls), not an arbitrary sample of the whole run. This is done
per-batch rather than pooling extremes across the whole dataset, since the
batches with the most severe drift would otherwise dominate which items
count as "extreme" and leave some batches with no anchor at all.

```{r fixedLabels}
score <- X_obs[, 1] + X_obs[, 2] # a simple separating projection
n_fixed_per_class_per_batch <- 4

fixed <- rep(0L, N)
for (b in 0:(B - 1)) {
  bidx <- which(batch_vec == b)
  neg_idx <- bidx[labels_true[bidx] == 0]
  pos_idx <- bidx[labels_true[bidx] == 1]
  extreme_neg <- neg_idx[order(score[neg_idx])][seq_len(min(n_fixed_per_class_per_batch, length(neg_idx)))]
  extreme_pos <- pos_idx[order(-score[pos_idx])][seq_len(min(n_fixed_per_class_per_batch, length(pos_idx)))]
  fixed[c(extreme_neg, extreme_pos)] <- 1L
}
cat("Fixed (confirmed-label) items per batch and class:\n")
print(table(batch = batch_vec[fixed == 1] + 1, class = labels_true[fixed == 1]))
```

## 2. Visualise the data, with and without the batch effect

```{r simViz}
plot_df <- data.frame(
  X1 = X_obs[, 1], X2 = X_obs[, 2],
  X1_true = X_true[, 1], X2_true = X_true[, 2],
  class = factor(labels_true, labels = c("negative", "positive")),
  batch = factor(batch_vec + 1),
  fixed = factor(fixed, labels = c("unlabelled", "confirmed"))
)

ggplot(plot_df, aes(X1, X2, colour = batch)) +
  geom_point(alpha = 0.6) +
  labs(title = "Observed data, coloured by batch", subtitle = "Batches visibly shifted/rescaled from each other, independently of collection order") +
  theme_minimal()

ggplot(plot_df, aes(X1, X2, colour = class, shape = fixed, alpha = fixed)) +
  geom_point() +
  scale_alpha_manual(values = c(unlabelled = 0.35, confirmed = 1)) +
  labs(
    title = "Observed data, coloured by true class",
    subtitle = "The two classes overlap far more in the noisiest batches (random per batch); confirmed labels are the extremes"
  ) +
  theme_minimal()

ggplot(plot_df, aes(X1_true, X2_true, colour = class)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "True batch-free data, coloured by class",
    subtitle = "The underlying biology: always well separated, regardless of batch"
  ) +
  theme_minimal()
```

```{r trueProportionViz}
prop_df <- data.frame(batch = factor(1:B), time = time_points, true_p1 = true_p1)
curve_grid <- data.frame(time = seq(0, end_t, length.out = 400))
curve_grid$true_p1 <- true_prevalence(curve_grid$time)

ggplot(prop_df, aes(time, true_p1)) +
  geom_line(data = curve_grid, colour = "grey50", linetype = "dashed") +
  geom_point(aes(colour = batch), size = 3) +
  geom_vline(xintercept = c(outbreak_t, wave2_t), linetype = "dotted", colour = "grey40") +
  labs(
    title = "True per-batch positive-class proportion",
    subtitle = "Baseline, outbreak + spread, intervention to a solution level, then a second wave (dotted lines: the two discontinuities)",
    x = "Batch collection time", y = "True P(positive)"
  ) +
  ylim(0, 1) +
  theme_minimal()
```

## 3. Fit three models

Identical in every respect (data, labels, `type = "MVN"`) except
`batch_weight_prior`. For `"gp"`, `batch_coordinates` is the actual
collection time (not batch index), and `sample_gp_hyperparameters = TRUE`
lets the length scale and marginal variance be estimated rather than fixed
- important here since there is no reason to assume any particular
smoothness in advance. Four chains per model, so agreement between chains
can be checked before trusting any of it (Section 4).

```{r fitModels}
n_chains <- 4
n_iter <- 1000
thin <- 15
n_burn0 <- 300

init_labels <- labels_true
init_labels[fixed == 0] <- sample(0:(K - 1), sum(fixed == 0), replace = TRUE)

out_global <- fitBatchMix(X_obs, n_chains, n_iter, thin, batch_vec + 1, "MVN",
  initial_labels = init_labels, fixed = fixed,
  control = batchmixControl(n_burn = n_burn0),
  batch_weight_prior = "global"
)

out_pp <- fitBatchMix(X_obs, n_chains, n_iter, thin, batch_vec + 1, "MVN",
  initial_labels = init_labels, fixed = fixed,
  control = batchmixControl(n_burn = n_burn0),
  batch_weight_prior = "partial_pooling"
)

out_gp <- fitBatchMix(X_obs, n_chains, n_iter, thin, batch_vec + 1, "MVN",
  initial_labels = init_labels, fixed = fixed,
  control = batchmixControl(n_burn = n_burn0),
  batch_weight_prior = "gp", batch_coordinates = time_points,
  sample_gp_hyperparameters = TRUE
)
```

## 4. MCMC diagnostics

Each fit already carries its own rank-normalized split-\(\hat{R}\)/ESS
report on the complete-data log-likelihood (`fitBatchMix()` computes this
automatically whenever `n_chains >= 2` - see `?assessConvergence`):

```{r printFits}
out_global
out_pp
out_gp
```

```{r diagPlots}
plotAcceptanceRates(out_global)
plotAcceptanceRates(out_pp)
plotAcceptanceRates(out_gp)

plotLikelihoods(out_pp)
plotLikelihoods(out_gp)
```

Per-batch weights under partial pooling and, especially, a Gaussian
process over irregular time points are a harder target than a single
global weight - a batch's positive-class weight is only weakly identified
by that batch's own (often heavily overlapping) data plus a handful of
confirmed labels, so it is worth checking convergence explicitly for this
quantity too, not only for the overall likelihood the reports above cover.
`rankNormalizedRhat()` (the same rank-normalized, folded diagnostic behind
the reports above) is exported and works directly on any raw sampled
quantity such as this one:

```{r rhat}
discontinuity_weight_rhat <- function(out, label, batch_idx, jump_name) {
  # This batch's positive-class weight: the first batch at/after a genuine
  # discontinuity, and so among the hardest for either of these two priors
  # to pin down. Not checked for "global": every batch is identical by
  # construction there, so a per-batch Rhat tests nothing beyond the
  # likelihood report above.
  w_b <- sapply(out, function(ch) ch$w_batch[batch_idx, 2, ])
  half <- seq(floor(nrow(w_b) / 2) + 1, nrow(w_b))
  cat(label, "Rhat, batch", batch_idx, paste0("(", jump_name, ")"), "positive-class weight:",
      round(rankNormalizedRhat(w_b[half, ])$rhat, 3), "\n")
}

discontinuity_weight_rhat(out_pp, "partial_pooling", outbreak_batch_idx, "outbreak")
discontinuity_weight_rhat(out_gp, "gp", outbreak_batch_idx, "outbreak")
discontinuity_weight_rhat(out_pp, "partial_pooling", wave2_batch_idx, "second wave")
discontinuity_weight_rhat(out_gp, "gp", wave2_batch_idx, "second wave")
```

If any diagnostic above is still well above the usual ~1.01 rule of thumb,
the right move is to run longer via `continueChains()` before trusting
anything downstream - not to explain it away.

```{r continueChains}
n_iter_new <- 5000

out_global <- continueChains(out_global, X_obs, fixed, batch_vec + 1, n_iter_new, keep_old_samples = TRUE)
out_pp <- continueChains(out_pp, X_obs, fixed, batch_vec + 1, n_iter_new, keep_old_samples = TRUE)
out_gp <- continueChains(out_gp, X_obs, fixed, batch_vec + 1, n_iter_new, keep_old_samples = TRUE)

out_global
out_pp
out_gp

discontinuity_weight_rhat(out_pp, "partial_pooling (continued)", outbreak_batch_idx, "outbreak")
discontinuity_weight_rhat(out_gp, "gp (continued)", outbreak_batch_idx, "outbreak")
discontinuity_weight_rhat(out_pp, "partial_pooling (continued)", wave2_batch_idx, "second wave")
discontinuity_weight_rhat(out_gp, "gp (continued)", wave2_batch_idx, "second wave")
```

```{r processChains}
n_iter_total <- n_iter + n_iter_new
burn_total <- floor(n_iter_total / 2)

proc_global <- processMCMCChains(out_global, burn_total)[[1]]
proc_pp <- processMCMCChains(out_pp, burn_total)[[1]]
proc_gp <- processMCMCChains(out_gp, burn_total)[[1]]
```

```{r rhatFinalEcho, echo=FALSE}
final_rhats <- c(
  global = attr(out_global, "convergence")$rhat,
  partial_pooling = attr(out_pp, "convergence")$rhat,
  gp = attr(out_gp, "convergence")$rhat
)
worst_model <- names(which.max(final_rhats))
```

R-hat generally drops after continuing, but is unlikely to be perfectly at
1 for every quantity in every one of the three fits: here, after
continuing, complete-likelihood R-hat is
`r round(final_rhats["global"], 2)` for `global`,
`r round(final_rhats["partial_pooling"], 2)` for `partial_pooling` and
`r round(final_rhats["gp"], 2)` for `gp` - `r worst_model` is the
highest of the three on this run. There is no guarantee any one of the
three is *always* hardest to pin down: `global` has the fewest parameters
but the poorest-fitting model (a systematically wrong model can itself
show up as between-chain disagreement, since different chains may settle
on different ways of being wrong); `gp`'s Gaussian process over `r B`
irregular time points, fit to data with two real discontinuities in what
it is modelling, has a structural reason to mix slowly regardless of fit
quality. Whichever is worst on a given run, the lesson is the same as
elsewhere in this package's vignettes: check R&#770; for the actual fit in
front of you rather than assuming a ranking from a previous run or a
different dataset. This vignette is kept to a single continuation call to keep
its build time reasonable; in real use, keep calling `continueChains()`
until every diagnostic of interest is comfortably below ~1.1, not just
the ones checked here.

## 5. Recovered per-batch proportions vs the truth

The plot the whole comparison hinges on: each method's posterior point
estimate of the positive-class proportion in every batch, against the true
epidemic curve from Section 2.

```{r weightRecoveryViz}
weight_recovery_df <- data.frame(
  batch = rep(1:B, 3),
  time = rep(time_points, 3),
  true_p1 = rep(true_p1, 3),
  # w_batch_est only exists for batch_weight_prior != "global" (there is no
  # per-batch weight to estimate under a single global weight) - the global
  # model's single positive-class weight is repeated across every batch
  # instead, since that is exactly what "global" means: one shared estimate
  # applied everywhere, regardless of batch.
  est_p1 = c(
    rep(median(proc_global$weights[, 2]), B),
    proc_pp$w_batch_est[, 2], proc_gp$w_batch_est[, 2]
  ),
  method = rep(c("global", "partial_pooling", "gp"), each = B)
)

ggplot(weight_recovery_df, aes(time, true_p1)) +
  geom_line(aes(y = true_p1), colour = "grey40", linetype = "dashed") +
  geom_point(aes(y = true_p1), colour = "grey20", shape = 4, size = 3, stroke = 1.2) +
  geom_line(aes(y = est_p1, colour = method)) +
  geom_point(aes(y = est_p1, colour = method), size = 2.5) +
  labs(
    title = "Recovered vs true per-batch positive-class proportion",
    subtitle = "Grey X = truth (also the dashed line); coloured points/lines = each method's posterior estimate",
    x = "Batch collection time", y = "P(positive)"
  ) +
  ylim(0, 1) +
  theme_minimal()
```

```{r weightRecoverySummary, echo=FALSE}
global_flat_est <- median(proc_global$weights[, 2])
pp_err <- abs(proc_pp$w_batch_est[, 2] - true_p1)
gp_err <- abs(proc_gp$w_batch_est[, 2] - true_p1)
worst_pp_batch <- which.max(pp_err)
worst_gp_batch <- which.max(gp_err)
near_jump <- function(b) {
  gap_outbreak <- abs(time_points[b] - outbreak_t)
  gap_wave2 <- abs(time_points[b] - wave2_t)
  if (min(gap_outbreak, gap_wave2) < 2) {
    paste0("close in time to the ", if (gap_outbreak < gap_wave2) "outbreak" else "second-wave", " discontinuity")
  } else {
    "not particularly close in time to either discontinuity - just a batch whose own data happened to be weakly informative (e.g. heavier assay noise, or an unlucky draw of confirmed labels)"
  }
}
```

`global` is flat by construction, at `r round(global_flat_est, 2)`
(roughly the pooled average prevalence across all twelve batches,
`r round(mean(true_p1), 2)`) - it has no way to represent either
discontinuity at all, so it is systematically wrong for most of the
window: too high during the baseline and solution periods, too low near
each peak. `partial_pooling` and `gp` both track the curve's ups and downs
far more closely - which is the main result: either is a real fix for what
`global` cannot do at all. Neither is perfect everywhere: `partial_pooling`
comes closest to failing at batch `r worst_pp_batch`
(t = `r round(time_points[worst_pp_batch], 1)`, true P(positive) =
`r round(true_p1[worst_pp_batch], 2)`, estimated
`r round(proc_pp$w_batch_est[worst_pp_batch, 2], 2)`), and `gp` at batch
`r worst_gp_batch` (t = `r round(time_points[worst_gp_batch], 1)`, true
P(positive) = `r round(true_p1[worst_gp_batch], 2)`, estimated
`r round(proc_gp$w_batch_est[worst_gp_batch, 2], 2)`) - batch
`r worst_pp_batch` is `r near_jump(worst_pp_batch)`, and batch
`r worst_gp_batch` is `r near_jump(worst_gp_batch)`. A batch immediately
next to a jump is the mechanistically expected hard case: both models'
information-sharing (shrinkage towards a common mean for
`partial_pooling`, towards nearby batches in time for `gp`) is exactly
what lets them represent a jump at all, and the same mechanism is what
pulls a batch right at the boundary towards its still-unshifted or
already-shifted neighbours instead of its own true value.

## 6. Batch-corrected data per method, vs ground truth

Each model's `inferred_dataset` (its posterior batch-corrected estimate)
coloured by *predicted* class, compared against the true batch-free data
coloured by *true* class from Section 2.

```{r correctedViz, fig.width=7, fig.height=5.5}
make_corrected_df <- function(proc, method) {
  data.frame(
    X1 = proc$inferred_dataset[, 1], X2 = proc$inferred_dataset[, 2],
    pred_class = factor(proc$pred - 1, labels = c("negative", "positive")),
    method = method
  )
}

corrected_df <- rbind(
  make_corrected_df(proc_global, "global"),
  make_corrected_df(proc_pp, "partial_pooling"),
  make_corrected_df(proc_gp, "gp")
)
truth_df <- data.frame(
  X1 = X_true[, 1], X2 = X_true[, 2],
  pred_class = factor(labels_true, labels = c("negative", "positive")),
  method = "truth"
)
combined_df <- rbind(corrected_df, truth_df)
combined_df$method <- factor(combined_df$method, levels = c("truth", "global", "partial_pooling", "gp"))

ggplot(combined_df, aes(X1, X2, colour = pred_class)) +
  geom_point(alpha = 0.6, size = 1.2) +
  facet_wrap(~method) +
  labs(
    title = "Batch-corrected data by method vs ground truth",
    subtitle = "Coloured by predicted class (truth panel: true class)"
  ) +
  theme_minimal()
```

## 7. Classification accuracy by batch

A single number per method hides exactly the batches this scenario is
designed to stress, so this breaks accuracy down by batch instead of just
reporting one overall figure.

```{r accuracy}
unlabelled <- fixed == 0

acc_by_batch <- function(proc, method) {
  pred <- proc$pred - 1L
  data.frame(
    batch = 1:B, time = time_points, method = method,
    accuracy = as.numeric(tapply(pred[unlabelled] == labels_true[unlabelled], batch_vec[unlabelled], mean))
  )
}

acc_df <- rbind(
  acc_by_batch(proc_global, "global"),
  acc_by_batch(proc_pp, "partial_pooling"),
  acc_by_batch(proc_gp, "gp")
)

ggplot(acc_df, aes(batch, accuracy, colour = method, group = method)) +
  geom_line() +
  geom_point(size = 2.5) +
  geom_vline(
    xintercept = c(outbreak_batch_idx, wave2_batch_idx) - 0.5,
    linetype = "dotted", colour = "grey50"
  ) +
  labs(
    title = "Classification accuracy on unlabelled items, by batch",
    subtitle = "Dotted lines mark the first batch at/after each discontinuity",
    x = "Batch", y = "Accuracy"
  ) +
  ylim(0, 1) +
  theme_minimal()

cat("Overall accuracy (unlabelled items):\n")
cat("  global:         ", round(mean((proc_global$pred - 1L)[unlabelled] == labels_true[unlabelled]), 3), "\n")
cat("  partial_pooling:", round(mean((proc_pp$pred - 1L)[unlabelled] == labels_true[unlabelled]), 3), "\n")
cat("  gp:             ", round(mean((proc_gp$pred - 1L)[unlabelled] == labels_true[unlabelled]), 3), "\n")
```

```{r accuracySummary, echo=FALSE}
high_prev_batches <- which(true_p1 > 0.4) # the peak/spread/second-wave batches
acc_range <- function(m, batches) round(range(acc_df$accuracy[acc_df$method == m & acc_df$batch %in% batches]), 2)
global_high <- acc_range("global", high_prev_batches)
pp_high <- acc_range("partial_pooling", high_prev_batches)
gp_high <- acc_range("gp", high_prev_batches)
mean_acc_by_batch <- tapply(acc_df$accuracy, acc_df$batch, mean)
hardest_batch <- as.integer(names(which.min(mean_acc_by_batch)))
```

`global`'s accuracy in the elevated-prevalence batches (`r paste(high_prev_batches, collapse = ", ")`,
i.e. those with true P(positive) above 0.4) is `r global_high[1]`-`r global_high[2]`
here: it keeps assuming a mostly-negative prior even where the true
proportion has flipped towards mostly-positive, so it systematically
under-calls the positive class exactly where it has become common.
`partial_pooling` (`r pp_high[1]`-`r pp_high[2]`) and `gp`
(`r gp_high[1]`-`r gp_high[2]`) both recover noticeably more accuracy in
those same batches - this is the practical payoff of Section 5's weight
recovery: getting the per-batch prior closer to right directly improves
the classifications that prior feeds into. Averaged across all three
methods, batch `r hardest_batch` (t = `r round(time_points[hardest_batch], 1)`)
is the hardest overall - it is `r near_jump(hardest_batch)` - the same
kind of case Section 5 flagged as hardest to estimate a weight for, for
the same reason: heavy class overlap and sitting right at the edge of a
regime change, which no choice of weight prior can fully fix once the raw
signal itself is this ambiguous.

## 8. Posterior predictive check

Do the three fitted models reproduce the one thing this vignette is
actually about - the *per-batch* positive-class proportion - not just get
the individual classifications right? This is the same idea as the
getting-started vignette's per-batch-mean posterior predictive check,
applied here to a per-batch proportion instead of a per-batch mean, and it
is a check the point estimates in Section 5 cannot provide on their own:
a point estimate can look reasonable by luck on a single fit, whereas a
predictive check asks whether the *fitted model's own generative process*
reproduces the observed jump at the outbreak batch, not just whether a
summary of it happens to land close by.

```{r postPCWeights, fig.width=7}
post_sims_global <- simulatePosteriorPredictive(getBestChain(out_global), batch_vec + 1, n_draws = 40, burn = burn_total)
post_sims_pp <- simulatePosteriorPredictive(getBestChain(out_pp), batch_vec + 1, n_draws = 40, burn = burn_total)
post_sims_gp <- simulatePosteriorPredictive(getBestChain(out_gp), batch_vec + 1, n_draws = 40, burn = burn_total)

# A replicate's "positive-class proportion" for the outbreak batch has no
# direct class label (simulatePosteriorPredictive() replicates X, not the
# allocations) - proxy it with the fraction of that batch's replicated
# column 1 values above the observed data's own midpoint between the two
# class means at that batch, which is a fixed, model-independent threshold
# applied identically to every replicate and to the observed data.
check_batch <- outbreak_batch_idx
check_batch_0idx <- check_batch - 1L # batch_vec's own 0-indexed convention
midpoint <- mean(mu_true[, 1]) + batch_shift_true[check_batch]
high_frac <- function(x_col, batch, b) mean(x_col[batch == b] > midpoint)
stat_name <- function(method) paste0("Batch ", check_batch, " (outbreak): frac(col 1 > midpoint) - ", method)

plotPredictiveCheck(X_obs, post_sims_global,
  style = "statistic", column = 1,
  statistic = function(x) high_frac(x, batch_vec, check_batch_0idx),
  statistic_name = stat_name("global")
)
plotPredictiveCheck(X_obs, post_sims_pp,
  style = "statistic", column = 1,
  statistic = function(x) high_frac(x, batch_vec, check_batch_0idx),
  statistic_name = stat_name("partial_pooling")
)
plotPredictiveCheck(X_obs, post_sims_gp,
  style = "statistic", column = 1,
  statistic = function(x) high_frac(x, batch_vec, check_batch_0idx),
  statistic_name = stat_name("gp")
)
```

`global`'s replicated fraction is generated under a *single* class weight
shared by every batch, so its replicates should cluster around the
pre-outbreak (low) rate rather than around this batch's actually-elevated
observed rate - a mismatch here is not a sign of a broken fit, it is
exactly the limitation Section 5 already showed numerically, now visible
as a predictive check instead of a point estimate. `partial_pooling` and
`gp` each have their own weight for this specific batch, so their
replicates should track the observed value far more closely - if either
did not, that would be a genuine warning sign about that fit specifically,
not just about the general limitation of a shared weight. The same check
could equally be run at the second-wave batch (`r wave2_batch_idx`); it is
not repeated here to keep this vignette's length reasonable.
