---
title: "batchmix workflow"
author: "Stephen Coleman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using BatchMixtureModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
set.seed(1)
```

## Introduction

This is the end-to-end workflow for `batchmix`: simulate data with a known
batch effect, confirm the batch effect is actually there (and check what the
data would look like without it), fit the semi-supervised model, check that
the MCMC chains have actually converged before trusting anything they say,
and finally check that the fitted model recovered the parameters used to
generate the data and removed the batch effect. Each step matters on its own
- a model fit from a chain that hasn't converged, or accepted without
checking whether the batch correction did anything, is not a result you can
trust.

```{r libraries, message=FALSE}
library(batchmix)
library(ggplot2)
```

## 1. Simulate data with a known batch effect

`generateBatchData()` returns both the `observed_data` (what you would
actually measure, batch effect included) and the `corrected_data` (the same
draws with the batch effect never applied) for the same items, which is what
makes it possible to check, on this simulation, whether a fitted model's
batch correction actually recovers the right thing rather than just
producing *some* transformation of the data.

```{r dataGen}
N <- 400
P <- 2
K <- 3
B <- 5

group_means <- c(0, 5, 10)
group_sds <- rep(1, K)

# A deliberately obvious batch effect: shifts of very different sizes, and
# scales further from 1 than you would want in practice, so that its effect
# on the raw data is unambiguous in the plots below. (The model's batch
# scale has support (1, Inf) - see S_loc in the package's C++ sources - so
# generateBatchData() requires batch_scale >= 1 too.)
batch_shift <- c(-4, -2, 0, 2, 4)
batch_scale <- c(1.0, 1.3, 1.6, 2.0, 2.6)

group_weights <- rep(1 / K, K)
batch_weights <- rep(1 / B, B)

my_data <- generateBatchData(
  N, P,
  group_means, group_sds,
  batch_shift, batch_scale,
  group_weights, batch_weights,
  type = "MVN"
)

X <- my_data$observed_data
X_true <- my_data$corrected_data
true_labels <- my_data$group_IDs
batch_vec <- my_data$batch_IDs
fixed <- my_data$fixed
```

`fixed` marks which items have an observed label (this simulation labels
about 20% of items, `frac_known` in `generateBatchData()`); the rest are
what the model has to cluster.

## 2. Confirm the batch effect is actually present

Before fitting anything, look at the raw data. Colouring the *same*
`observed_data` by batch and then by true group shows whether the batch
effect is visible at all, and whether it's large enough to matter relative
to the group structure; colouring the batch-free `corrected_data` by group
shows what the data would look like if the batch effect were successfully
removed - the target the model fit is aiming for.

```{r batchEffectViz}
plot_df <- data.frame(
  X1 = X[, 1], X2 = X[, 2],
  X1_true = X_true[, 1], X2_true = X_true[, 2],
  batch = factor(batch_vec),
  group = factor(true_labels)
)

# Observed data, coloured by batch: if there were no batch effect, batches
# would be scattered on top of each other rather than forming separated
# clouds.
ggplot(plot_df, aes(X1, X2, colour = batch)) +
  geom_point(alpha = 0.7) +
  labs(title = "Observed data, coloured by batch", subtitle = "Batches visibly offset/rescaled: the batch effect is present") +
  theme_minimal()

# The same observed data, coloured by the true group instead: batch noise
# make the group structure noisier/harder to see than in the corrected data
# below.
ggplot(plot_df, aes(X1, X2, colour = group)) +
  geom_point(alpha = 0.7) +
  labs(title = "Observed data, coloured by true group", subtitle = "Batch effect blurs the group separation") +
  theme_minimal()

# The batch-free ground truth: what a perfect batch correction would recover.
ggplot(plot_df, aes(X1_true, X2_true, colour = group)) +
  geom_point(alpha = 0.7) +
  labs(title = "Batch-free ground truth (corrected_data), coloured by true group", subtitle = "Target for the batch-corrected posterior estimate") +
  theme_minimal()
```

## 3. Fit the model

We run several chains in parallel so we can check they agree (Section 4) -
a single chain can look perfectly converged and still be wrong if it's
stuck in a poor part of the space. `auto_tune = TRUE` (the default) adapts
every Metropolis-Hastings proposal window automatically during `n_burn`
iterations, rather than requiring hand-tuned windows up front.

```{r runMCMCChains}
R <- 1500
thin <- 25
n_chains <- 4

alpha <- 1
initial_labels <- generateInitialLabels(alpha, K, fixed, true_labels)

mcmc_output <- runMCMCChains(
  X, n_chains, R, thin, batch_vec, "MVN",
  initial_labels = initial_labels,
  fixed = fixed,
  auto_tune = TRUE,
  n_burn = floor(R / 2)
)
```

## 4. MCMC diagnostics

**Acceptance rates.** Metropolis-Hastings proposals accepted far too often
(close to 1) or almost never (close to 0) both indicate a poorly-tuned
chain that isn't exploring properly; `auto_tune` aims most blocks at
~0.234-0.44 automatically, but it's still worth checking nothing pinned at
an extreme.

```{r plotAcceptanceRatesEarly}
plotAcceptanceRates(mcmc_output)
```

**Likelihood traces.** `complete_likelihood` (label-dependent, most relevant
to the clustering itself) and `observed_likelihood` (label-free, more
relevant to density estimation) should each look like they've settled down
to fluctuating around a stable level, with the chains agreeing with each
other, not still trending or visibly separated.

```{r likelihood}
plotLikelihoods(mcmc_output)
```

**Multi-chain agreement (Gelman-Rubin R-hat).** A visual check is a good
start but is easy to kid yourself with; the standard quantitative check is
the potential scale reduction factor (Gelman & Rubin, 1992, *Statistical
Science* 7(4)): compare the variance of a quantity *within* each chain to
its variance pooling *across* chains, for the second half of each chain
(to let early-chain differences burn off). Values close to 1 (rule of
thumb: below ~1.1) indicate the chains agree; values well above 1 mean at
least one chain hasn't converged to the same place as the others.

```{r rhat}
gelmanRubinRhat <- function(chain_matrix) {
  # chain_matrix: iterations x chains, already restricted to the second
  # half of each chain.
  n <- nrow(chain_matrix)
  m <- ncol(chain_matrix)
  chain_means <- colMeans(chain_matrix)
  grand_mean <- mean(chain_means)

  # Between-chain variance (scaled as in Gelman & Rubin, 1992).
  B <- n / (m - 1) * sum((chain_means - grand_mean)^2)
  # Within-chain variance.
  W <- mean(apply(chain_matrix, 2, var))

  var_hat <- ((n - 1) / n) * W + B / n
  sqrt(var_hat / W)
}

complete_lkl <- sapply(mcmc_output, function(chain) chain$complete_likelihood)
second_half <- seq(floor(nrow(complete_lkl) / 2) + 1, nrow(complete_lkl))

cat("R-hat for complete_likelihood:", round(gelmanRubinRhat(complete_lkl[second_half, ]), 3), "\n")

# The same check for a specific, interpretable parameter: the first batch's
# shift on the first feature.
batch1_shift_p1 <- sapply(mcmc_output, function(chain) chain$batch_shift[1, 1, ])
cat("R-hat for batch 1's shift (feature 1):", round(gelmanRubinRhat(batch1_shift_p1[second_half, ]), 3), "\n")
```

If either diagnostic looks bad - trending likelihoods, R-hat well above
1.1, or acceptance rates stuck near 0 or 1 - the right move is to run
longer before trusting anything downstream, via `continueChains()`:

```{r continueChains}
R_new <- 6000

new_output <- continueChains(
  mcmc_output, X, fixed, batch_vec, R_new,
  keep_old_samples = TRUE
)

plotLikelihoods(new_output)

rhat_after <- gelmanRubinRhat(
  sapply(new_output, function(chain) chain$complete_likelihood)[
    seq(floor(R_new / thin / 2) + 1, floor((R + R_new) / thin)),
  ]
)
cat("R-hat for complete_likelihood after continuing:", round(rhat_after, 3), "\n")
```

R-hat has moved closer to 1 than the first check, but this vignette is kept
deliberately short so it builds quickly - in real use, keep calling
`continueChains()` (each call picks up exactly where the last left off)
until R-hat is comfortably below ~1.1 for every quantity you care about,
not just the one or two checked here, before trusting any point estimate
from `processMCMCChains()` below.

## 5. Process the chains

```{r processChains}
burn <- floor((R + R_new) / 2)
processed_samples <- processMCMCChains(new_output, burn)
chain_used <- processed_samples[[1]]
```

## 6. Parameter recovery

Because this is simulated data, we know the true group means and batch
shifts, so we can check directly whether the model recovered them (up to
label switching between the sampler's cluster indices and the simulator's
group indices, and up to the usual mean/shift additive confound when
neither is anchored - here we just match each sampled cluster to its
closest true group mean for a readable plot).

```{r paramRecovery}
est_means <- chain_used$mean_est # P x K
match_order <- apply(est_means, 2, function(col) which.min(abs(col[1] - group_means)))

recovery_df <- data.frame(
  true_mean = group_means[match_order],
  est_mean = est_means[1, ],
  cluster = factor(seq_len(K))
)

ggplot(recovery_df, aes(true_mean, est_mean, colour = cluster)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3) +
  labs(
    title = "Cluster mean recovery (feature 1)",
    x = "True group mean", y = "Posterior point estimate"
  ) +
  theme_minimal()

shift_recovery_df <- data.frame(
  true_shift = batch_shift,
  est_shift = chain_used$shift_est[1, ],
  batch = factor(seq_len(B))
)

ggplot(shift_recovery_df, aes(true_shift, est_shift, colour = batch)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3) +
  labs(
    title = "Batch shift recovery (feature 1)",
    x = "True batch shift", y = "Posterior point estimate"
  ) +
  theme_minimal()
```

## 7. Did the batch correction work?

The real test: compare the model's inferred batch-corrected data
(`inferred_dataset`, the posterior estimate) against the simulator's
`corrected_data` ground truth from Section 2, not just against the raw
observed data.

```{r finalViz}
pc_raw <- prcomp(X, scale. = TRUE)
pc_corrected <- prcomp(chain_used$inferred_dataset, scale. = TRUE)
pc_truth <- prcomp(X_true, scale. = TRUE)

viz_df <- data.frame(
  PC1_raw = pc_raw$x[, 1], PC2_raw = pc_raw$x[, 2],
  PC1_est = pc_corrected$x[, 1], PC2_est = pc_corrected$x[, 2],
  PC1_truth = pc_truth$x[, 1], PC2_truth = pc_truth$x[, 2],
  pred_labels = factor(chain_used$pred),
  true_labels = factor(true_labels),
  prob = chain_used$prob,
  batch = factor(batch_vec)
)

ggplot(viz_df, aes(PC1_raw, PC2_raw, colour = batch)) +
  geom_point(alpha = 0.7) +
  labs(title = "Raw observed data (PCA)", subtitle = "Batch effect still visible") +
  theme_minimal()

ggplot(viz_df, aes(PC1_est, PC2_est, colour = pred_labels, alpha = prob)) +
  geom_point() +
  labs(title = "Model's batch-corrected estimate (PCA)", subtitle = "Coloured by predicted cluster") +
  theme_minimal()

ggplot(viz_df, aes(PC1_truth, PC2_truth, colour = true_labels)) +
  geom_point(alpha = 0.7) +
  labs(title = "True batch-free data (PCA)", subtitle = "What the estimate above should resemble") +
  theme_minimal()

test_inds <- which(fixed == 0)
accuracy <- sum(true_labels[test_inds] == chain_used$pred[test_inds]) / length(test_inds)
cat("Classification accuracy on unlabelled items:", round(accuracy, 3), "\n")
```

If the batch-corrected PCA plot looks like the true batch-free plot (same
group separation, no leftover batch structure) rather than like the raw
data, the correction worked; the classification accuracy on the unlabelled
items gives a single-number summary of the same thing for the class
structure specifically.
