---
title: "Getting started: an end-to-end batchmix workflow"
author: "batchmix"
output: rmarkdown::html_vignette
description: |
  A complete, self-contained batchmix analysis: simulate data with a known
  batch effect, check the model's priors before fitting, fit and diagnose
  multiple MCMC chains, check the fitted model's predictions, and verify
  that both the clustering and the batch correction recovered the truth.
vignette: >
  %\VignetteIndexEntry{Getting started: an end-to-end batchmix workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
set.seed(1)
```

## Introduction

Fitting a model and reading off a point estimate of the clustering is only
the middle of a Bayesian analysis. This vignette works through the full
loop on a single simulated dataset:

1. **Simulate** data with a known, deliberately visible batch effect, so
   every later step can be checked against ground truth.
2. **Prior predictive check**: before the model sees the real
   clustering/batch structure, do datasets simulated from its priors alone
   resemble plausible data?
3. **Fit** several independent MCMC chains.
4. **MCMC diagnostics**: acceptance rates, likelihood traces, and
   rank-normalized split-\(\hat{R}\)/effective sample size across chains,
   checked before the chains are trusted.
5. **Posterior predictive check**: does the fitted model reproduce the
   features of the observed data, overall and for the per-batch statistics
   the batch correction is meant to equalise?
6. **Point estimate and parameter recovery**: does the fitted model recover
   the cluster means and batch effects used to generate the data?
7. **Did the batch correction work?**: comparing the model's inferred
   batch-corrected data against the simulator's batch-free ground truth.

Steps 2 and 5 are standard Bayesian workflow, not specific to this
package: Gelman, Carlin, Stern, Dunson, Vehtari & Rubin (2013), *Bayesian
Data Analysis*, 3rd ed. ("BDA3"), devote Chapter 6 to this pair of checks;
Gelman, Vehtari, Simpson, Margossian, Carpenter, Yao, Kennedy, Gabry,
Bürkner & Modrák (2020), "Bayesian Workflow" (arXiv:2011.01808), Section 2,
frame both as the two ends of the modelling loop - simulate before
fitting, criticise after. Step 5 matters particularly here because a batch
effect is an unobservable confound by construction: real data has no
batch-free ground truth to check the correction against directly, so
predictive checks are among the few tools available for detecting a
correction that is systematically wrong.

```{r libraries, message=FALSE}
library(batchmix)
library(ggplot2)
```

## 1. Simulate data with a known batch effect

`generateBatchData()` returns both the `observed_data` (what would be
measured, batch effect included) and the `corrected_data` (the same
draws with the batch effect never applied) for the same items - this is
what makes it possible to check, on this simulation, whether a fitted
model's batch correction recovers the right thing, rather than just
producing *some* transformation of the data.

```{r dataGen}
N <- 400
P <- 2
K <- 3
B <- 5

group_means <- c(0, 5, 10)
group_sds <- rep(1, K)

# A deliberately obvious batch effect: shifts of very different sizes, and
# scales further from 1 than would be typical in practice, so that its
# effect on the raw data is unambiguous in the plots below. (The model's
# batch scale has support (1, Inf), so generateBatchData() requires
# batch_scale >= 1 too.)
batch_shift <- c(-4, -2, 0, 2, 4)
batch_scale <- c(1.0, 1.3, 1.6, 2.0, 2.6)

group_weights <- rep(1 / K, K)
batch_weights <- rep(1 / B, B)

my_data <- generateBatchData(
  N, P,
  group_means, group_sds,
  batch_shift, batch_scale,
  group_weights, batch_weights,
  type = "MVN"
)

X <- my_data$observed_data
X_true <- my_data$corrected_data
true_labels <- my_data$group_IDs
batch_vec <- my_data$batch_IDs
fixed <- my_data$fixed
```

`fixed` marks which items have an observed label (this simulation labels
about 20% of items, `frac_known` in `generateBatchData()`); the rest are
what the model has to cluster.

## 2. Confirm the batch effect is present

Before fitting anything, look at the raw data. Colouring the *same*
`observed_data` by batch and then by true group shows whether the batch
effect is visible at all, and whether it is large enough to matter relative
to the group structure; colouring the batch-free `corrected_data` by group
shows what the data would look like if the batch effect were successfully
removed - the target the model fit is aiming for.

```{r batchEffectViz}
plot_df <- data.frame(
  X1 = X[, 1], X2 = X[, 2],
  X1_true = X_true[, 1], X2_true = X_true[, 2],
  batch = factor(batch_vec),
  group = factor(true_labels)
)

# Observed data, coloured by batch: if there were no batch effect, batches
# would be scattered on top of each other rather than forming separated
# clouds.
ggplot(plot_df, aes(X1, X2, colour = batch)) +
  geom_point(alpha = 0.7) +
  labs(title = "Observed data, coloured by batch", subtitle = "Batches visibly offset/rescaled: the batch effect is present") +
  theme_minimal()

# The same observed data, coloured by the true group instead: batch noise
# makes the group structure noisier/harder to see than in the corrected
# data below.
ggplot(plot_df, aes(X1, X2, colour = group)) +
  geom_point(alpha = 0.7) +
  labs(title = "Observed data, coloured by true group", subtitle = "Batch effect blurs the group separation") +
  theme_minimal()

# The batch-free ground truth: what a perfect batch correction would recover.
ggplot(plot_df, aes(X1_true, X2_true, colour = group)) +
  geom_point(alpha = 0.7) +
  labs(title = "Batch-free ground truth (corrected_data), coloured by true group", subtitle = "Target for the batch-corrected posterior estimate") +
  theme_minimal()
```

## 3. Prior predictive check

Simulate datasets purely from the model's priors (empirical-Bayes priors
centred/scaled from `X`'s own summary statistics, exactly as the real
fit's constructor computes them - see `?simulatePriorPredictive`) and look
at whether they resemble the kind of data expected here. This is
deliberately done at the *default* hyperparameters - the ones every call
to `fitBatchMix()`/`runBatchMix()` uses unless overridden - since those are
the ones being checked.

```{r priorPC}
prior_sims <- simulatePriorPredictive(
  X, batch_vec,
  K = K, type = "MVN",
  n_datasets = 30
)

plotPredictiveCheck(X, prior_sims, style = "density", column = 1)
```

The observed density (black) should sit comfortably *inside* the envelope
of prior-simulated densities (blue) - not because the prior should match
the data closely (it has not seen the clustering yet, so it should not),
but because the *scale* of variation it is prepared to generate should be
plausible for data of this kind. A prior predictive envelope that is
wildly wider or narrower than anything like the observed data is a sign to
reconsider the prior (or its empirical-Bayes hyperparameters,
e.g. `rho`/`theta`/`m_scale`) before ever looking at a fit.

A scalar summary makes the same check quantitative - the classical
prior/posterior predictive check (BDA3 Eq. 6.9): compute a test statistic
on the observed data and on every replicate, and see where the observed
value falls in the replicated distribution.

```{r priorPCStat}
plotPredictiveCheck(X, prior_sims,
  style = "statistic", column = 1,
  statistic = stats::sd, statistic_name = "SD(column 1)"
)
```

## 4. Fit multiple chains

Several independent chains are run so that they can be checked against
each other (Step 5) - a single chain can look perfectly converged and
still be wrong if it is stuck in a poor part of the space, which is a real
risk for mixture models specifically (see `?generateInitialLabels`).
`auto_tune = TRUE` (the default, bundled with the other sampler-tuning
settings into `control` - see `?batchmixControl`) adapts every
Metropolis-Hastings proposal window automatically during burn-in, rather
than requiring hand-tuned windows up front.

```{r fitBatchMix}
n_iter <- 2000
thin <- 20
n_chains <- 4

alpha <- 1
initial_labels <- generateInitialLabels(alpha, K, fixed, true_labels)

mcmc_output <- fitBatchMix(
  X, n_chains, n_iter, thin, batch_vec, "MVN",
  initial_labels = initial_labels,
  fixed = fixed
)
```

`fitBatchMix()` returns a `batchmix_fit_list` and, since `n_chains >= 2`,
has already computed the cross-chain convergence diagnostics used in Step
5 below (shown above in the message, and reachable at any time by printing
the fit itself):

```{r printFit}
mcmc_output
```

## 5. MCMC diagnostics

**Acceptance rates.** Metropolis-Hastings proposals accepted far too often
(close to 1) or almost never (close to 0) both indicate a poorly-tuned
chain that is not exploring properly; `auto_tune` aims most blocks at
~0.234-0.44 automatically, but it is still worth checking nothing pinned
at an extreme.

```{r plotAcceptanceRatesEarly}
plotAcceptanceRates(mcmc_output)
```

**Likelihood traces.** `complete_likelihood` (label-dependent, most
relevant to the clustering itself) and `observed_likelihood` (label-free,
more relevant to density estimation) should each look like they have
settled down to fluctuating around a stable level, with the chains
agreeing with each other, not still trending or visibly separated.

```{r likelihood}
plotLikelihoods(mcmc_output)
```

**Multi-chain agreement (\(\hat{R}\)).** A visual check is a good start
but is easy to be fooled by. The report printed above already gives the
quantitative check: rank-normalized, folded, split-\(\hat{R}\) (Vehtari,
Gelman, Simpson, Carpenter & Bürkner, 2021, *Bayesian Analysis* 16(2)) -
the diagnostic that superseded the classical Gelman & Rubin (1992)
potential scale reduction factor, because the classical version can fail
to catch chains that agree on location but disagree on scale, or in the
tails. It is computed on the **complete-data log-likelihood trace**, not
raw parameters like `batch_shift`, because mixture models are only
identified up to a permutation of the cluster labels - a naive
\(\hat{R}\) on a raw parameter would confuse harmless label-switching
between chains with genuine non-convergence (see `?assessConvergence`).
Values close to 1 (rule of thumb: below ~1.01) indicate the chains agree;
values well above 1 mean at least one chain has not converged to the same
place as the others.

If that looks bad - trending likelihoods, \(\hat{R}\) well above 1.01, or
acceptance rates stuck near 0 or 1 - simply running longer is often *not*
enough by itself: with well-separated clusters, a chain that starts from a
poor random partition can get stuck in a persistently worse-fitting local
optimum that single-site Metropolis-Hastings essentially never escapes, no
matter how long it runs (a high \(\hat{R}\) that does not improve with a
longer chain is the signature of this). `runBatchMix()`/`fitBatchMix()`
mitigate this automatically for a fully unsupervised fit by seeding from a
k-means partition of the data instead of a random draw; here the fit is
semi-supervised, so the labelled points already anchor cluster identity
and this failure mode is much less likely, but it is still worth
increasing `n_chains` (more chances at least one lands well) as well as
chain length. When it *is* just a matter of needing more samples, extend
every chain with `continueChains()`, which picks up exactly where each
chain left off and recomputes convergence on the extended chains:

```{r continueChains}
n_iter_new <- 4000
max_extra_rounds <- 3
round_i <- 0

new_output <- continueChains(
  mcmc_output, X, fixed, batch_vec, n_iter_new,
  keep_old_samples = TRUE
)

# Actually do what the advice above says, rather than just asserting it:
# keep extending every chain until Rhat is comfortable or a bounded number
# of extra rounds have been tried (a real analysis would not bound this,
# but an unbounded loop is unsuitable for a vignette that has to finish
# building in reasonable time). Each round is wrapped in tryCatch(): an
# extremely long-running chain can occasionally hit a numerically singular
# covariance update (a rare edge case, not the usual outcome) - if that
# happens, this stops the loop and keeps the last successful state rather
# than losing everything already sampled.
extension_failed <- FALSE
repeat {
  if (attr(new_output, "convergence")$rhat <= 1.05 || round_i >= max_extra_rounds) {
    break
  }
  next_output <- tryCatch(
    continueChains(new_output, X, fixed, batch_vec, n_iter_new, keep_old_samples = TRUE),
    error = function(e) {
      message("continueChains() round ", round_i + 1, " failed (", conditionMessage(e), "); keeping the last successful state.")
      NULL
    }
  )
  if (is.null(next_output)) {
    extension_failed <- TRUE
    break
  }
  new_output <- next_output
  round_i <- round_i + 1
}

new_output
```

```{r continueChainsEcho, echo=FALSE}
final_rhat <- attr(new_output, "convergence")$rhat
final_chain_bic <- attr(new_output, "convergence")$chain_bic
bic_spread <- diff(range(final_chain_bic))
```

The first `continueChains()` call above (2000 to 6000 total iterations)
already ran unconditionally; the loop then attempted `r round_i`
*additional* extension(s) beyond that because R&#770; was still above 1.05.
`r if (extension_failed) "One further extension attempt hit a numerically singular covariance update partway through - a rare edge case in the underlying sampler, not the usual outcome - and was safely discarded by the tryCatch() above, keeping the chains at their last successful state rather than losing everything already sampled." else "No extension attempt failed."`
R&#770; after all of this is `r round(final_rhat, 2)`. In real use, keep
calling `continueChains()` (without the artificial round limit used here)
until R&#770; is comfortably below ~1.01 before trusting any point
estimate; this vignette stops early to keep its build time reasonable, and
whatever value R&#770; has actually reached is what the rest of this
section relies on - not an assumption that it has hit 1.

If R&#770; is still above ~1.01 at this point, the per-chain mean BIC
(`r round(final_chain_bic, 1)`, spread `r round(bic_spread, 1)`) is worth
checking directly: R&#770; on the complete-data log-likelihood aggregates
*all* chains' disagreement into a single number, whereas the per-chain BIC
shows *which* chains disagree and by how much. A single point estimate
below is only reasonable to trust here because `getBestChain()`
specifically extracts the chain with the best BIC rather than pooling
across chains that may not agree - it sidesteps the disagreement R&#770;
is flagging rather than resolving it, which is a meaningfully weaker
guarantee than "all chains agree" and is exactly why R&#770; is still
reported rather than hidden once a best chain has been picked.

## 6. Posterior predictive check

Now simulate replicate datasets from the *fitted* posterior - each
replicate uses one retained MCMC iteration's actual drawn cluster means,
covariances, batch shifts/scales and cluster allocations (see
`?simulatePosteriorPredictive`; no prior formula is re-derived here, so
this cannot drift out of sync with the sampler itself). `getBestChain()`
picks the chain with the best (highest post-burn-in mean) BIC rather than
an arbitrary one - reasonable as a single representative chain even while
R&#770; is not yet at 1, for the reason discussed just above (it sidesteps
cross-chain disagreement rather than requiring it to be resolved first).

```{r bestChain}
mcmc_out <- getBestChain(new_output)
summary(mcmc_out)
```

```{r postPC}
# mcmc_out$n_iter is the chain's own actual total iteration count, however
# many continueChains() rounds it took above - using it here (rather than
# a fixed n_iter + n_iter_new) stays correct regardless of how many rounds
# that loop actually ran.
burn <- floor(mcmc_out$n_iter / 2)

post_sims <- simulatePosteriorPredictive(
  mcmc_out, batch_vec,
  n_draws = 30, burn = burn
)

plotPredictiveCheck(X, post_sims, style = "density", column = 1)
```

Unlike the prior predictive check, the posterior predictive envelope
*should* now track the observed density closely - the model has seen the
data. A visible, systematic mismatch here (not just replicate-to-replicate
noise) is evidence the model is missing something structural: too few
clusters, the wrong density family (see the covariance-model vignette for
`type = "MVT"` if the observed data has heavier tails than every
replicate), or a batch effect the correction is not fully removing.

That last case is the one specific to this package, and worth checking
directly: compute a per-batch statistic (e.g. each batch's mean) on the
observed data and on every replicate. If the model's batch correction is
capturing the real batch effect, the *observed* per-batch means should
land inside the *replicated* per-batch means' spread about as often as
chance would suggest - not systematically outside it.

```{r postPCBatch}
batch1_mean <- function(x_col, batch, b) mean(x_col[batch == b])

plotPredictiveCheck(X, post_sims,
  style = "statistic", column = 1,
  statistic = function(x) batch1_mean(x, batch_vec, batch_vec[1]),
  statistic_name = paste0("Mean(column 1, batch ", batch_vec[1], ")")
)
```

`plotBatchCorrection()` complements this numerical check with the more
immediately readable picture: the raw data against the point-estimate
batch-corrected data, coloured by batch. A working correction pulls the
batch colours together from "Raw" to "Batch-corrected" while the cluster
point-clouds stay recognisable.

```{r batchCorrectionPlot}
plotBatchCorrection(mcmc_out, X, batch_vec, burn = burn)
```

## 7. Point estimate and parameter recovery

`processMCMCChain()` applies the burn-in and collapses the retained
samples down to point estimates - `mean_est`/`shift_est` for the cluster
means and batch shifts, `pred`/`prob` for the predicted class and its
probability, and `inferred_dataset` for the point-estimate batch-corrected
data.

```{r processChains}
chain_used <- processMCMCChain(mcmc_out, burn)
```

Because this is simulated data, the true group means and batch shifts are
known, so they can be checked directly against the fit (up to label
switching between the sampler's cluster indices and the simulator's group
indices, and up to the usual mean/shift additive confound when neither is
anchored - here each sampled cluster is simply matched to its closest true
group mean for a readable plot).

```{r paramRecovery}
est_means <- chain_used$mean_est # P x K
match_order <- apply(est_means, 2, function(col) which.min(abs(col[1] - group_means)))

recovery_df <- data.frame(
  true_mean = group_means[match_order],
  est_mean = est_means[1, ],
  cluster = factor(seq_len(K))
)

ggplot(recovery_df, aes(true_mean, est_mean, colour = cluster)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3) +
  labs(
    title = "Cluster mean recovery (feature 1)",
    x = "True group mean", y = "Posterior point estimate"
  ) +
  theme_minimal()

shift_recovery_df <- data.frame(
  true_shift = batch_shift,
  est_shift = chain_used$shift_est[1, ],
  batch = factor(seq_len(B))
)

ggplot(shift_recovery_df, aes(true_shift, est_shift, colour = batch)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3) +
  labs(
    title = "Batch shift recovery (feature 1)",
    x = "True batch shift", y = "Posterior point estimate"
  ) +
  theme_minimal()
```

## 8. Did the batch correction work?

The real test: compare the model's inferred batch-corrected data
(`inferred_dataset`, the posterior estimate) against the simulator's
`corrected_data` ground truth from Step 2, not just against the raw
observed data.

```{r finalViz}
pc_raw <- prcomp(X, scale. = TRUE)
pc_corrected <- prcomp(chain_used$inferred_dataset, scale. = TRUE)
pc_truth <- prcomp(X_true, scale. = TRUE)

viz_df <- data.frame(
  PC1_raw = pc_raw$x[, 1], PC2_raw = pc_raw$x[, 2],
  PC1_est = pc_corrected$x[, 1], PC2_est = pc_corrected$x[, 2],
  PC1_truth = pc_truth$x[, 1], PC2_truth = pc_truth$x[, 2],
  pred_labels = factor(chain_used$pred),
  true_labels = factor(true_labels),
  prob = chain_used$prob,
  batch = factor(batch_vec)
)

ggplot(viz_df, aes(PC1_raw, PC2_raw, colour = batch)) +
  geom_point(alpha = 0.7) +
  labs(title = "Raw observed data (PCA)", subtitle = "Batch effect still visible") +
  theme_minimal()

ggplot(viz_df, aes(PC1_est, PC2_est, colour = pred_labels, alpha = prob)) +
  geom_point() +
  labs(title = "Model's batch-corrected estimate (PCA)", subtitle = "Coloured by predicted cluster") +
  theme_minimal()

ggplot(viz_df, aes(PC1_truth, PC2_truth, colour = true_labels)) +
  geom_point(alpha = 0.7) +
  labs(title = "True batch-free data (PCA)", subtitle = "What the estimate above should resemble") +
  theme_minimal()

test_inds <- which(fixed == 0)
accuracy <- sum(true_labels[test_inds] == chain_used$pred[test_inds]) / length(test_inds)
cat("Classification accuracy on unlabelled items:", round(accuracy, 3), "\n")
```

If the batch-corrected PCA plot looks like the true batch-free plot (same
group separation, no leftover batch structure) rather than like the raw
data, the correction worked; the classification accuracy on the unlabelled
items gives a single-number summary of the same thing for the class
structure specifically.

## Summary

- Simulate with a known ground truth first, so every later step - priors,
  fit, and correction - can be checked against something.
- Prior predictive check: are the *default* priors capable of generating
  plausible-looking data at this `K`/`B`, before any fitting happens?
- Fit several chains, not one, and check acceptance rates, likelihood
  traces and \(\hat{R}\)/ESS (`print()`/`summary()` on the fit report all
  three) before trusting anything downstream; extend with
  `continueChains()` rather than restarting when more samples are needed.
- Posterior predictive check: does the *fitted* model reproduce the
  features of the observed data - overall, and specifically for
  the per-batch statistics the batch correction is meant to equalise?
  `plotBatchCorrection()` gives the same check as a picture.
- Only then take a point estimate (`processMCMCChain()`/`getBestChain()`,
  or pool every chain with `predictFromMultipleChains()`), and check it
  against ground truth when it is available. If R&#770; is still elevated
  even after continuing, `getBestChain()`'s single best-BIC chain
  sidesteps the cross-chain disagreement rather than resolving it - a
  meaningfully weaker guarantee than every chain actually agreeing, worth
  keeping in mind rather than treating a printed R&#770; as a box to tick
  once a point estimate has been produced regardless of its value.

Neither predictive check *proves* the model correct - as with all
posterior predictive checks, a good result is consistent with (not proof
of) an adequate model, and both checks are as sensitive as the test
statistics/columns chosen to look at (BDA3 Section 6.3 discusses this
limitation directly). But routinely looking, in both directions, is what
turns a single point-estimate fit into a Bayesian workflow - and every
other vignette in this package follows the same loop, adapted to whichever
density (`type`), covariance prior, or batch-weight structure it is
demonstrating.
