---
title: "Correlated batch effects on class proportions: global, partial-pooling and GP weight priors"
author: "batchmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correlated batch effects on class proportions: global, partial-pooling and GP weight priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5)
```

## Purpose

`batch_weight_prior` controls how the class/cluster mixing weights are
allowed to vary across batches: `"global"` ties every batch to one shared
weight vector, `"partial_pooling"` lets each batch have its own weight with
no assumed order or structure (only exchangeable shrinkage towards a common
mean), and `"gp"` places a Gaussian process over a batch coordinate (e.g.
collection time), so batches close together are expected to have similar
proportions. The other vignettes cover the mean/covariance side of batch
correction; this one is specifically about the *proportions*, using a
scenario designed so the three choices actually disagree with each other.

**Scenario.** Six assay batches (e.g. an ELISA seropositivity panel) are
collected at irregular points across a two-year window. Two things drift
with time, for different reasons:

- The assay itself drifts smoothly - reagent/instrument aging shifts and
  spreads the raw readout further from its true value the longer a batch
  is run after the assay was calibrated. This erodes how separable the two
  classes look in the *raw* data, even though nothing about the underlying
  biology has changed.
- The true proportion of positives does **not** drift smoothly - it is flat
  and low for the first four batches, then jumps abruptly at batch 5 (an
  outbreak) and stays high at batch 6. This is a step change, not a trend,
  and it is what the three `batch_weight_prior` choices disagree about:
  `"global"` cannot represent it at all, `"partial_pooling"` can represent
  it immediately with no assumption about batch order, and `"gp"` assumes
  nearby batches have similar proportions - an assumption a genuine step
  change violates right at the boundary, making it the hardest of the
  three priors to get exactly right here, even though it is the only one
  built to use real collection-time information at all.

```{r libs, message=FALSE}
library(batchmix)
library(ggplot2)
```

## 1. Simulate the data

`generateBatchData()` cannot produce this scenario (it has no notion of a
time-varying batch effect or batch-specific class proportions), so the data
is simulated directly. Batch collection times are irregular - drawn at
random over a 24-unit window, not evenly spaced - since that is what
`batch_weight_prior = "gp"` is actually built for and it is worth stressing
that the model uses genuine inter-batch distances, not just batch order.

```{r simSetup}
set.seed(2024)

B <- 6
time_points <- sort(round(runif(B, 0, 24), 2))
cat("Batch collection times:", time_points, "\n")
cat("Gaps between consecutive batches:", round(diff(time_points), 2), "\n")
```

With this particular random draw, batches 4 and 5 land only a fraction of
a unit apart in time - essentially back-to-back - which makes the outbreak
an even sharper test for `batch_weight_prior = "gp"` than the scenario
already is by design: two batches that are about as close together as any
pair in this dataset turn out to have almost the most different true
proportions of any pair (0.15 vs 0.65). A GP has no time-based warning
that anything is about to change.

```{r simSetupContinued}
K <- 2 # seronegative (0) / seropositive (1)
P <- 2 # two correlated assay readouts

# True, batch-free class means: fixed, well-separated - the biology itself
# does not change over time, only what is measured does.
mu_true <- rbind(c(0, 0), c(5, 5))
class_sd <- 1.0

# Smooth assay drift: both the shift and the scale grow with time. Scale
# growth (not shift) is what actually erodes separability - an additive
# shift moves both classes by the same amount and so does not by itself
# bring them closer together.
drift_rate <- 0.09
scale_rate <- 0.085
batch_shift_true <- drift_rate * time_points
batch_scale_true <- 1 + scale_rate * time_points # S_loc has support [1, Inf)

# True per-batch prevalence: flat and low for batches 1-4, a step change at
# batch 5 that persists into batch 6.
true_p1 <- c(0.15, 0.10, 0.20, 0.15, 0.65, 0.75)
```

```{r simGenerate}
n_per_batch <- 100
N <- n_per_batch * B

X_obs <- matrix(0, N, P)
X_true <- matrix(0, N, P)
labels_true <- integer(N)
batch_vec <- integer(N)

row_i <- 1
for (b in seq_len(B)) {
  n1 <- rbinom(1, n_per_batch, true_p1[b])
  n0 <- n_per_batch - n1
  cls <- c(rep(0, n0), rep(1, n1))
  for (k in cls) {
    z <- rnorm(P, mean = mu_true[k + 1, ], sd = class_sd)
    X_true[row_i, ] <- z
    X_obs[row_i, ] <- (z - mu_true[k + 1, ]) * batch_scale_true[b] +
      mu_true[k + 1, ] + batch_shift_true[b]
    labels_true[row_i] <- k
    batch_vec[row_i] <- b - 1
    row_i <- row_i + 1
  }
}
```

**Semi-supervised labels: the extremes, not a random sample.** Rather than
labelling a random subset (`generateBatchData()`'s usual `frac_known`),
here the "confirmed" items are the most extreme observed values within
each true class, per batch - the items a lab would realistically have
independent confirmatory testing for (e.g. a panel's positive/negative
controls), not an arbitrary sample of the whole run. This is done
per-batch rather than pooling extremes across the whole dataset, since the
batches with the most severe drift would otherwise dominate which items
count as "extreme" and leave some batches with no anchor at all.

```{r fixedLabels}
score <- X_obs[, 1] + X_obs[, 2] # a simple separating projection
n_fixed_per_class_per_batch <- 4

fixed <- rep(0L, N)
for (b in 0:(B - 1)) {
  bidx <- which(batch_vec == b)
  neg_idx <- bidx[labels_true[bidx] == 0]
  pos_idx <- bidx[labels_true[bidx] == 1]
  extreme_neg <- neg_idx[order(score[neg_idx])][seq_len(min(n_fixed_per_class_per_batch, length(neg_idx)))]
  extreme_pos <- pos_idx[order(-score[pos_idx])][seq_len(min(n_fixed_per_class_per_batch, length(pos_idx)))]
  fixed[c(extreme_neg, extreme_pos)] <- 1L
}
cat("Fixed (confirmed-label) items per batch and class:\n")
print(table(batch = batch_vec[fixed == 1] + 1, class = labels_true[fixed == 1]))
```

## 2. Visualise the data, with and without the batch effect

```{r simViz}
plot_df <- data.frame(
  X1 = X_obs[, 1], X2 = X_obs[, 2],
  X1_true = X_true[, 1], X2_true = X_true[, 2],
  class = factor(labels_true, labels = c("negative", "positive")),
  batch = factor(batch_vec + 1),
  fixed = factor(fixed, labels = c("unlabelled", "confirmed"))
)

ggplot(plot_df, aes(X1, X2, colour = batch)) +
  geom_point(alpha = 0.6) +
  labs(title = "Observed data, coloured by batch", subtitle = "Later batches visibly shifted and more spread out") +
  theme_minimal()

ggplot(plot_df, aes(X1, X2, colour = class, shape = fixed, alpha = fixed)) +
  geom_point() +
  scale_alpha_manual(values = c(unlabelled = 0.35, confirmed = 1)) +
  labs(
    title = "Observed data, coloured by true class",
    subtitle = "The two classes overlap far more in later (more drifted) batches; confirmed labels are the extremes"
  ) +
  theme_minimal()

ggplot(plot_df, aes(X1_true, X2_true, colour = class)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "True batch-free data, coloured by class",
    subtitle = "The underlying biology: always well separated, regardless of batch"
  ) +
  theme_minimal()
```

```{r trueProportionViz}
prop_df <- data.frame(batch = factor(1:B), time = time_points, true_p1 = true_p1)
ggplot(prop_df, aes(time, true_p1)) +
  geom_line(linetype = "dashed", colour = "grey50") +
  geom_point(aes(colour = batch), size = 3) +
  labs(
    title = "True per-batch positive-class proportion",
    subtitle = "Flat through batches 1-4, a step change at batch 5 that persists into batch 6",
    x = "Batch collection time", y = "True P(positive)"
  ) +
  ylim(0, 1) +
  theme_minimal()
```

## 3. Fit three models

Identical in every respect (data, labels, `type = "MVN"`) except
`batch_weight_prior`. For `"gp"`, `batch_coordinates` is the actual
collection time (not batch index), and `sample_gp_hyperparameters = TRUE`
lets the length scale and marginal variance be estimated rather than fixed
- important here since there is no reason to assume any particular
smoothness in advance. Four chains per model, so agreement between chains
can be checked before trusting any of it (Section 4).

```{r fitModels}
n_chains <- 4
R0 <- 6000
thin <- 15
n_burn0 <- 3000

init_labels <- labels_true
init_labels[fixed == 0] <- sample(0:(K - 1), sum(fixed == 0), replace = TRUE)

out_global <- fitBatchMix(X_obs, n_chains, R0, thin, batch_vec + 1, "MVN",
  initial_labels = init_labels, fixed = fixed,
  auto_tune = TRUE, n_burn = n_burn0,
  batch_weight_prior = "global"
)

out_pp <- fitBatchMix(X_obs, n_chains, R0, thin, batch_vec + 1, "MVN",
  initial_labels = init_labels, fixed = fixed,
  auto_tune = TRUE, n_burn = n_burn0,
  batch_weight_prior = "partial_pooling"
)

out_gp <- fitBatchMix(X_obs, n_chains, R0, thin, batch_vec + 1, "MVN",
  initial_labels = init_labels, fixed = fixed,
  auto_tune = TRUE, n_burn = n_burn0,
  batch_weight_prior = "gp", batch_coordinates = time_points,
  sample_gp_hyperparameters = TRUE
)
```

## 4. MCMC diagnostics

```{r diagPlots}
plotAcceptanceRates(out_global)
plotAcceptanceRates(out_pp)
plotAcceptanceRates(out_gp)

plotLikelihoods(out_pp)
plotLikelihoods(out_gp)
```

Per-batch weights under partial pooling and, especially, a Gaussian
process over irregular time points are a harder target than a single
global weight - a batch's positive-class weight is only weakly identified
by that batch's own (often heavily overlapping) data plus a handful of
confirmed labels, so it is worth checking convergence explicitly for this
quantity rather than only for the overall likelihood.

```{r rhat}
gelmanRubinRhat <- function(chain_matrix) {
  # chain_matrix: iterations x chains, restricted to the second half of
  # each chain.
  n <- nrow(chain_matrix)
  m <- ncol(chain_matrix)
  chain_means <- colMeans(chain_matrix)
  grand_mean <- mean(chain_means)
  B_var <- n / (m - 1) * sum((chain_means - grand_mean)^2)
  W_var <- mean(apply(chain_matrix, 2, var))
  sqrt((((n - 1) / n) * W_var + B_var / n) / W_var)
}

rhat_summary <- function(out, label, check_batch_weight = TRUE) {
  lkl <- sapply(out, function(ch) ch$complete_likelihood)
  half <- seq(floor(nrow(lkl) / 2) + 1, nrow(lkl))
  cat(label, "R-hat, complete_likelihood:", round(gelmanRubinRhat(lkl[half, ]), 3), "\n")
  if (check_batch_weight) {
    # Batch 5's positive-class weight: the batch right at the discontinuity,
    # and so the hardest one for either of these two priors to pin down.
    # (Not checked for "global": every batch is identical by construction,
    # so a per-batch R-hat there tests nothing beyond the likelihood check.)
    w5 <- sapply(out, function(ch) ch$w_batch[5, 2, ])
    cat(label, "R-hat, batch 5 positive-class weight:", round(gelmanRubinRhat(w5[half, ]), 3), "\n")
  }
}

rhat_summary(out_global, "global", check_batch_weight = FALSE)
rhat_summary(out_pp, "partial_pooling")
rhat_summary(out_gp, "gp")
```

If any of these are still well above the usual ~1.1 rule of thumb, the
right move is to run longer via `continueChains()` before trusting
anything downstream - not to explain it away.

```{r continueChains}
R_new <- 18000

out_global <- continueChains(out_global, X_obs, fixed, batch_vec + 1, R_new, keep_old_samples = TRUE)
out_pp <- continueChains(out_pp, X_obs, fixed, batch_vec + 1, R_new, keep_old_samples = TRUE)
out_gp <- continueChains(out_gp, X_obs, fixed, batch_vec + 1, R_new, keep_old_samples = TRUE)

rhat_summary(out_global, "global (continued)", check_batch_weight = FALSE)
rhat_summary(out_pp, "partial_pooling (continued)")
rhat_summary(out_gp, "gp (continued)")
```

```{r processChains}
R_total <- R0 + R_new
burn_total <- floor(R_total / 2)

proc_global <- processMCMCChains(out_global, burn_total)[[1]]
proc_pp <- processMCMCChains(out_pp, burn_total)[[1]]
proc_gp <- processMCMCChains(out_gp, burn_total)[[1]]
```

R-hat drops sharply after continuing, but is unlikely to be perfectly at 1
for every quantity in every one of the three fits - in particular, `gp`'s
complete-likelihood R-hat commonly stays the highest of the three even
after continuing, which is exactly what the discussion above predicts: a
Gaussian process over six irregular time points, fit to data with a real
discontinuity in what it is modelling, is the hardest of the three targets
to pin down. This vignette is kept to a single continuation call to keep
its build time reasonable; in real use, keep calling `continueChains()`
until every diagnostic you care about is comfortably below ~1.1, not just
the ones checked here.

## 5. Recovered per-batch proportions vs the truth

The plot the whole comparison hinges on: each method's posterior point
estimate of the positive-class proportion in every batch, against the true
step function from Section 2.

```{r weightRecoveryViz}
weight_recovery_df <- data.frame(
  batch = rep(1:B, 3),
  time = rep(time_points, 3),
  true_p1 = rep(true_p1, 3),
  # w_batch_est only exists for batch_weight_prior != "global" (there is no
  # per-batch weight to estimate under a single global weight) - the global
  # model's single positive-class weight is repeated across every batch
  # instead, since that is exactly what "global" means: one shared estimate
  # applied everywhere, regardless of batch.
  est_p1 = c(
    rep(median(proc_global$weights[, 2]), B),
    proc_pp$w_batch_est[, 2], proc_gp$w_batch_est[, 2]
  ),
  method = rep(c("global", "partial_pooling", "gp"), each = B)
)

ggplot(weight_recovery_df, aes(time, true_p1)) +
  geom_line(aes(y = true_p1), colour = "grey40", linetype = "dashed") +
  geom_point(aes(y = true_p1), colour = "grey20", shape = 4, size = 3, stroke = 1.2) +
  geom_line(aes(y = est_p1, colour = method)) +
  geom_point(aes(y = est_p1, colour = method), size = 2.5) +
  labs(
    title = "Recovered vs true per-batch positive-class proportion",
    subtitle = "Grey X = truth (also the dashed line); coloured points/lines = each method's posterior estimate",
    x = "Batch collection time", y = "P(positive)"
  ) +
  ylim(0, 1) +
  theme_minimal()
```

`global` is flat by construction, at roughly the pooled average prevalence
across all six batches (~0.35) - it has no way to represent batch 5's jump
at all, so it is badly wrong in both directions: too high for the
low-prevalence batches 1-4, too low for the high-prevalence batches 5-6.
`partial_pooling` and `gp` both track the step correctly - low through
batches 1-4, high at 5-6 - which is the main result: either is a real fix
for what `global` cannot do at all. Where they are not perfect, both are
imperfect in the same place and the same direction: batch 4, immediately
before the jump, is over-estimated by both (its true proportion is 0.15,
but both land closer to 0.28-0.30). That is not a coincidence - batch 4
has the most accumulated assay drift of any of the low-prevalence batches
(heaviest class overlap, so the weakest per-batch signal) and sits right
at the edge of the regime change, so both models' information-sharing
mechanism (shrinkage towards a common mean for `partial_pooling`, towards
nearby batches for `gp`) pulls its estimate towards the now-elevated
batches next to it. A hierarchical model borrowing strength across batches
is exactly what makes it able to represent the jump at all; the same
mechanism is what makes the batch right at the boundary the hardest one
to pin down precisely.

## 6. Batch-corrected data per method, vs ground truth

Each model's `inferred_dataset` (its posterior batch-corrected estimate)
coloured by *predicted* class, compared against the true batch-free data
coloured by *true* class from Section 2.

```{r correctedViz, fig.width=7, fig.height=5.5}
make_corrected_df <- function(proc, method) {
  data.frame(
    X1 = proc$inferred_dataset[, 1], X2 = proc$inferred_dataset[, 2],
    pred_class = factor(proc$pred - 1, labels = c("negative", "positive")),
    method = method
  )
}

corrected_df <- rbind(
  make_corrected_df(proc_global, "global"),
  make_corrected_df(proc_pp, "partial_pooling"),
  make_corrected_df(proc_gp, "gp")
)
truth_df <- data.frame(
  X1 = X_true[, 1], X2 = X_true[, 2],
  pred_class = factor(labels_true, labels = c("negative", "positive")),
  method = "truth"
)
combined_df <- rbind(corrected_df, truth_df)
combined_df$method <- factor(combined_df$method, levels = c("truth", "global", "partial_pooling", "gp"))

ggplot(combined_df, aes(X1, X2, colour = pred_class)) +
  geom_point(alpha = 0.6, size = 1.2) +
  facet_wrap(~method) +
  labs(
    title = "Batch-corrected data by method vs ground truth",
    subtitle = "Coloured by predicted class (truth panel: true class)"
  ) +
  theme_minimal()
```

## 7. Classification accuracy by batch

A single number per method hides exactly the batches this scenario is
designed to stress, so this breaks accuracy down by batch instead of just
reporting one overall figure.

```{r accuracy}
unlabelled <- fixed == 0

acc_by_batch <- function(proc, method) {
  pred <- proc$pred - 1L
  data.frame(
    batch = 1:B, time = time_points, method = method,
    accuracy = as.numeric(tapply(pred[unlabelled] == labels_true[unlabelled], batch_vec[unlabelled], mean))
  )
}

acc_df <- rbind(
  acc_by_batch(proc_global, "global"),
  acc_by_batch(proc_pp, "partial_pooling"),
  acc_by_batch(proc_gp, "gp")
)

ggplot(acc_df, aes(batch, accuracy, colour = method, group = method)) +
  geom_line() +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 4.5, linetype = "dotted", colour = "grey50") +
  labs(
    title = "Classification accuracy on unlabelled items, by batch",
    subtitle = "Dotted line marks the outbreak (batch 5 onward)",
    x = "Batch", y = "Accuracy"
  ) +
  ylim(0, 1) +
  theme_minimal()

cat("Overall accuracy (unlabelled items):\n")
cat("  global:         ", round(mean((proc_global$pred - 1L)[unlabelled] == labels_true[unlabelled]), 3), "\n")
cat("  partial_pooling:", round(mean((proc_pp$pred - 1L)[unlabelled] == labels_true[unlabelled]), 3), "\n")
cat("  gp:             ", round(mean((proc_gp$pred - 1L)[unlabelled] == labels_true[unlabelled]), 3), "\n")
```

`global`'s accuracy falls away from batch 4 onward and is at its worst in
batches 5-6 (~0.84-0.88 here): it keeps assuming the pre-outbreak
(mostly-negative) prior even after the true proportion has flipped towards
mostly-positive, so it systematically under-calls the positive class
exactly where it has become common. `partial_pooling` and `gp` both
recover noticeably more accuracy in those same two batches (~0.90-0.92) -
this is the practical payoff of Section 5's weight recovery: getting the
per-batch prior closer to right directly improves the classifications that
prior feeds into. All three methods are roughly tied, and all three at
their worst, on batch 4 - the same batch that Section 5 flagged as the
hardest to estimate a weight for, for the same reason: heavy class overlap
and sitting right at the edge of the regime change, which no choice of
weight prior can fully fix once the raw signal itself is this ambiguous.
