---
title: "Prior and posterior predictive checks: a Bayesian workflow"
author: "batchmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prior and posterior predictive checks: a Bayesian workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
set.seed(42)
```

## Purpose

Fitting the model and reading off a point estimate of the clustering is
only the middle of a Bayesian analysis, not the whole of it. This vignette
walks through the two checks that bracket it, using
`simulatePriorPredictive()` and `simulatePosteriorPredictive()`:

1. **Prior predictive check** (before the model ever sees the real
   clustering/batch structure): do datasets simulated purely from the prior
   look like the kind of thing we expect to see? If not, the prior - not
   yet the data - is the thing to reconsider.
2. **Posterior predictive check** (after fitting): does the *fitted* model
   generate datasets that reproduce the features of the data we actually
   observed? This is especially important here, because batch effects are
   an *unobservable* confound by construction - we never get to see the
   "true" batch-corrected values to check the correction against directly,
   so predictive checks are one of the few tools available to catch a
   correction that is systematically wrong.

This is standard practice, not specific to this package: Gelman, Carlin,
Stern, Dunson, Vehtari & Rubin (2013), *Bayesian Data Analysis*, 3rd ed.
("BDA3"), devote Chapter 6 to exactly this pair of checks; McElreath (2020),
*Statistical Rethinking*, 2nd ed., introduces the same idea from single
parameters in Section 3.3 up through multilevel/hierarchical models (the
closest analogue to this package's batch structure) in Chapters 12-14;
Gelman, Vehtari, Simpson, Margossian, Carpenter, Yao, Kennedy, Gabry,
Bürkner & Modrák (2020), "Bayesian Workflow" (arXiv:2011.01808), Section 2,
frame both checks as the two ends of the modelling loop - simulate before
fitting, criticise after.

## Simulate data with a known batch effect

```{r simData}
library(batchmix)
library(ggplot2)

N <- 300
P <- 2
K <- 2
B <- 4

group_means <- c(0, 4)
group_std_devs <- c(1, 1)
batch_shift <- rnorm(B, mean = 0.5, sd = 1.0)
batch_scale <- rep(1.1, B) + exp(rnorm(B, mean = -1.0, sd = 0.25))

sim <- generateBatchData(
  N, P,
  group_means, group_std_devs,
  batch_shift, batch_scale,
  group_weights = rep(1 / K, K),
  batch_weights = rep(1 / B, B)
)

X <- sim$observed_data
X_true <- sim$corrected_data
batch_vec <- sim$batch_IDs
labels_true <- sim$group_IDs

ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], batch = factor(batch_vec)), aes(X1, X2, colour = batch)) +
  geom_point(alpha = 0.6) +
  labs(title = "Simulated data", subtitle = "Batch effects present") +
  theme_minimal()

ggplot(data.frame(X1 = X_true[, 1], X2 = X_true[, 2], batch = factor(batch_vec)), aes(X1, X2, colour = batch)) +
  geom_point(alpha = 0.6) +
  labs(title = "Simulated data", subtitle = "Batch effects removed") +
  theme_minimal()
```

## 1. Prior predictive check

Before fitting anything, simulate datasets purely from the model's priors
(empirical-Bayes priors centred/scaled from `X`'s own summary statistics,
exactly as the real fit's constructor computes them - see
`?simulatePriorPredictive`) and look at whether they resemble the kind of
data we're prepared to see. This is deliberately done at the *default*
hyperparameters - the ones every call to `runBatchMix()` uses unless
overridden - since those are the ones actually being checked.

```{r priorPC}
prior_sims <- simulatePriorPredictive(
  X, batch_vec,
  K = K, type = "MVN",
  n_datasets = 30
)

plotPredictiveCheck(X, prior_sims, style = "density", column = 1)
```

The observed density (black) should sit comfortably *inside* the envelope
of prior-simulated densities (blue) - not because the prior should match
the data closely (it hasn't seen the clustering yet, so it shouldn't), but
because the *scale* of variation it's prepared to generate should be
plausible for data of this kind. A prior predictive envelope that is wildly
wider or narrower than anything like the observed data is a sign to
reconsider the prior (or its empirical-Bayes hyperparameters,
e.g. `rho`/`theta`/`m_scale`) before ever looking at the fit.

A scalar summary makes the same check quantitative - the classical
prior/posterior predictive check (BDA3 Eq. 6.9): compute a test statistic
on the observed data and on every replicate, and see where the observed
value falls in the replicated distribution.

```{r priorPCStat}
plotPredictiveCheck(X, prior_sims,
  style = "statistic", column = 1,
  statistic = stats::sd, statistic_name = "SD(column 1)"
)
```

## Fit multiple chains and check convergence

Run several independent chains rather than one. `fitBatchMix()`
automatically reports the rank-normalized, folded, split-\(\hat{R}\)
(Vehtari, Gelman, Simpson, Carpenter & Bürkner, 2021, "Rank-normalization,
folding, and localization: An improved \(\hat{R}\) for assessing
convergence of MCMC", *Bayesian Analysis* 16(2)) and bulk/tail effective
sample size across the chains, and identifies the chain with the best
(highest post-burn-in mean) BIC.

The diagnostic is computed on the **complete-data log-likelihood trace**,
not the raw cluster means/covariances, deliberately: mixture models are
only identified up to a permutation of the cluster labels, so a naive
\(\hat{R}\) on e.g. `mu` would confuse harmless label-switching between
chains with genuine non-convergence. The likelihood trace is invariant to
that permutation and is the safe default target - see `?assessConvergence`.

```{r fitChains}
n_iter <- 10000
thin <- 20

chains <- fitBatchMix(
  X, n_chains = 4, n_iter, thin, batch_vec, "MVN",
  K_max = K,
  verbose = TRUE
)
```

`Rhat` values comfortably above 1.01 mean these chains have not mixed. If
the Rhat above stays poor even after increasing `n_iter` by several times,
that is a sign it is *not* simply a matter of needing a longer run: printing
each chain's mean complete-data
log-likelihood shows the chains split into two persistent groups
(`sapply(chains, function(ch) mean(tail(ch$complete_likelihood, 100)))`),
one noticeably worse than the other, all the way through the chain, not a
chain that starts worse and catches up. That is a chain stuck in a
genuinely different, worse-fitting local optimum from the start - a real
risk for well-separated clusters under single-site Metropolis-Hastings,
distinct from (though related to) label-switching, since escaping the wrong
optimum needs a large, coordinated jump this sampler essentially never
proposes. `generateInitialLabels()`'s k-means-seeded default (used
automatically above, since no `initial_labels` were given) reduces this
risk relative to a random start but does not eliminate it, and running the
same chain for even longer does not help once it is stuck - more chains
(more chances at least one lands in the better optimum) is the more
reliable fix, which is exactly why fitting several chains and checking
Rhat, not fitting one chain and trusting it, is the workflow this vignette
demonstrates.

Once Rhat/ESS look good, either pool every chain (via
`predictFromMultipleChains()`, which does not use any of this - it pools
unconditionally, which is only safe/sensible once every chain has actually
converged to the same thing) or, when a single representative chain is
wanted instead (e.g. to inspect directly, or to `continueChain()`), take
the one identified as best by BIC - which is also the more defensible
choice **when Rhat is still poor**, since averaging a good chain's
parameter estimate with a stuck chain's would produce a point estimate
that represents neither:

```{r bestChain}
mcmc_out <- getBestChain(chains)
cat("Best chain BIC (post burn-in mean):", attr(mcmc_out, "convergence")$chain_bic, "\n")
```

## 2. Posterior predictive check

Now simulate replicate datasets from the *fitted* posterior - each
replicate uses one retained MCMC iteration's actual drawn cluster means,
covariances, batch shifts/scales and cluster allocations (see
`?simulatePosteriorPredictive`; no prior formula is re-derived here, so
this cannot drift out of sync with the sampler itself).

```{r postPC}
burn <- n_iter / 2

post_sims <- simulatePosteriorPredictive(
  mcmc_out, batch_vec,
  n_draws = 30, burn = burn
)

plotPredictiveCheck(X, post_sims, style = "density", column = 1)
```

Unlike the prior predictive check, the posterior predictive envelope
*should* now track the observed density closely - the model has seen the
data. A visible, systematic mismatch here (not just replicate-to-replicate
noise) is evidence the model is missing something structural: too few
clusters, the wrong density family (compare against `type = "MVT"` - see
the `mvn_vs_mvt` vignette - if the observed data has heavier tails than
every replicate), or a batch effect the correction isn't fully removing.

That last case is the one specific to this package, and worth checking
directly: compute a per-batch statistic (e.g. each batch's mean) on the
observed data and on every replicate. If the model's batch correction is
capturing the real batch effect, the *observed* per-batch means should
land inside the *replicated* per-batch means' spread about as often as
chance would suggest - not systematically outside it.

```{r postPCBatch}
batch1_mean <- function(x_col, batch, b) mean(x_col[batch == b])

plotPredictiveCheck(X, post_sims,
  style = "statistic", column = 1,
  statistic = function(x) batch1_mean(x, batch_vec, batch_vec[1]),
  statistic_name = paste0("Mean(column 1, batch ", batch_vec[1], ")")
)
```

## Visualise the batch correction directly

`plotBatchCorrection()` complements the numerical check above with the more
immediately readable picture: the raw data against the point-estimate
batch-corrected data, coloured by batch. A working correction pulls the
batch colours together from "Raw" to "Batch-corrected" while the two
cluster point-clouds stay recognisable.

```{r batchCorrectionPlot}
plotBatchCorrection(mcmc_out, X, batch_vec, burn = burn)
```

## Summary

- Fit several chains, not one, and check `Rhat`/ESS before trusting
  anything downstream; use the chain with the best BIC (`getBestChain()`)
  when a single representative chain is wanted, or pool all of them with
  `predictFromMultipleChains()`.
- Prior predictive check: are the *default* priors capable of generating
  plausible-looking data at this `K`/`B`, before any fitting happens?
- Posterior predictive check: does the *fitted* model reproduce the
  features of the data we actually observed - overall, and specifically
  for the per-batch statistics that the batch correction is meant to
  equalise? `plotBatchCorrection()` gives the same check as a picture.

Neither check *proves* the model correct - as with all posterior predictive
checks, a good result is consistent with (not proof of) an adequate model,
and both checks are as sensitive as the test statistics/columns chosen to
look at (BDA3 Section 6.3 discusses this limitation directly). But routinely
looking, in both directions, is what turns a single point-estimate fit into
a Bayesian workflow.
