---
title: "Choosing a cluster covariance model: robustness and correlation"
author: "batchmix"
output: rmarkdown::html_vignette
description: |
  Comparing the three continuous-data covariance models batchmix offers -
  MVN, MVT and MVN_LKJ - against data with a known ground truth: comparing
  candidate cluster counts via BICM, does MVT's robustness matter for
  heavy-tailed data, and does the LKJ prior recover a known correlation
  structure.
vignette: >
  %\VignetteIndexEntry{Choosing a cluster covariance model: robustness and correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
```

## Purpose

batchmix offers three continuous-data cluster covariance models, sharing
the same batch-correction machinery and directly comparable as fits to the
same data:

- `type = "MVN"`: multivariate normal, Inverse-Wishart covariance prior.
  The default, and the right choice unless one of the two reasons below
  applies.
- `type = "MVT"`: multivariate $t$, with a per-cluster degrees-of-freedom
  parameter learned from the data. Robust to heavy-tailed/outlier-
  contaminated data, where MVN's Gaussian likelihood is not just imprecise
  but actively misleading (Part B below).
- `type = "MVN_LKJ"`: multivariate normal with an LKJ prior on the cluster
  *correlation* matrix and independent log-normal priors on the marginal
  *scales*, decoupling two things an Inverse-Wishart prior couples
  together (Part C below).

This vignette validates each against data with a **known ground truth**
from simulation - for checking whether a sampler recovers a *known*
generative model, this is the strictest possible test: the alternative (a
real dataset with expert-assigned labels) has its own label noise, so a
sampler could look "wrong" for reasons that have nothing to do with the
implementation.

## The hierarchical model

For item $i = 1, \dots, N$, with cluster label $c_i \in \{1,\dots,K\}$ and
batch label $b_i \in \{1,\dots,B\}$:

$$
x_i \mid c_i = k,\ b_i = b \ \sim\
\begin{cases}
\mathcal{N}\!\left(\mu_k + m_b,\ S_b \bullet \Sigma_k\right) & \texttt{"MVN"} \\[4pt]
t_{\nu_k}\!\left(\mu_k + m_b,\ S_b \bullet \Sigma_k\right) & \texttt{"MVT"} \\[4pt]
\mathcal{N}\!\left(\mu_k + m_b,\ S_b \bullet (\Delta_k R_k \Delta_k)\right) & \texttt{"MVN\_LKJ"}
\end{cases}
$$

where $S_b = \mathrm{diag}(S_{b,1}, \dots, S_{b,P})$ is a per-batch,
per-feature multiplicative scale, $m_b$ is a per-batch additive shift, and
$\bullet$ rescales only the *diagonal* of the cluster covariance,
$(S_b \bullet \Sigma_k)_{pp} = S_{b,p}\,\Sigma_{k,pp}$ and
$(S_b \bullet \Sigma_k)_{pq} = \Sigma_{k,pq}$ for $p \neq q$ - i.e. the batch
scale inflates each feature's marginal variance but leaves the cluster's raw
covariances between features untouched (this is deliberate: batch effects
are modelled as per-feature measurement noise, not a change in how features
correlate). This is *not* the same as the full congruence transform
$S_b^{1/2} \Sigma_k S_b^{1/2}$ except in the special case where $\Sigma_k$ is
itself diagonal (no correlation between features). Equivalently,
$S_b \bullet \Sigma_k = \Sigma_k + \mathrm{diag}\big((S_b - 1)\,
\mathrm{diag}(\Sigma_k)\big)$ (this needs $S_{b,p} \ge 1$, which the
`S_loc = 1` support of the batch-scale prior guarantees): an item is its
batch-free signal $c \sim \mathcal{N}(0, \Sigma_k)$ plus independent
per-feature batch noise, on top of the shift $m_b$.

The batch-corrected data returned in `batch_corrected_data` (and averaged
in `inferred_dataset`) follow directly from that decomposition: they are the
posterior mean of the batch-free signal,
$\mu_k + \Sigma_k (S_b \bullet \Sigma_k)^{-1}(x - \mu_k - m_b)$
(plus the interaction term when it is in use), which is the exact
inverse of the fitted batch model for any $\Sigma_k$. It is a denoising
estimate, not a change of variables: within-cluster deviations are shrunk
towards $\mu_k$ by $\Sigma_k (S_b \bullet \Sigma_k)^{-1}$, so the corrected
data are less variable than the batch-free signal itself.

For `"MVN"`/`"MVT"`, $\mu_k, \Sigma_k$
are the cluster mean and covariance with $\Sigma_k \sim \mathrm{IW}(\nu,
\Psi)$. `"MVN_LKJ"` decomposes the
covariance as $\Delta_k = \mathrm{diag}(\sigma_{k,1}, \dots, \sigma_{k,P})$
with **independent** priors on the correlation matrix and the marginal
scales,

$$
R_k \sim \mathrm{LKJ}(\eta), \qquad \log \sigma_{k,p} \sim \mathcal{N}(\beta, \xi^2),
$$

which lets a weakly-informative guess at the scales be stated without
implying anything about the correlations, or vice versa - something an
Inverse-Wishart prior makes hard, since it couples the two (Barnard,
McCulloch & Meng, 2000, *Statistica Sinica*; Tokuda, Goodrich, Van
Mechelen, Gelman & Meng, 2011, "Visualizing Distributions of Covariance
Matrices"). $\mathrm{LKJ}(\eta)$ (Lewandowski, Kurowicka & Joe, 2009, *J.
Multivariate Analysis*) has density $p(R \mid \eta) \propto \det(R)^{\eta -
1}$: $\eta = 1$ is uniform over the space of valid correlation matrices,
$\eta > 1$ concentrates mass towards the identity (shrinks correlations
towards 0) - it is the standard modern choice for this decomposition (Stan
and brms use it the same way).

The remaining priors are shared by every type:

$$
\mu_k \mid \Sigma_k \sim \mathcal{N}\!\left(\xi, \Sigma_k / \kappa\right), \qquad
m_b \sim \mathcal{N}(0, \delta^2 \lambda^2), \quad \lambda^2 \sim \mathrm{InvGamma}(3, 1), \qquad
S_{b,p} - 1 \sim \mathrm{InvGamma}(\rho, \theta),
$$

with $\nu_k - \nu_{\min} \sim \mathrm{Gamma}(\psi, \chi)$ additionally for
`"MVT"`. Two of these are *partially pooled* by default. The batch scales
$S_{b,p}$ are exchangeable draws from a common InvGamma whose concentration
is estimated, $\rho - 2 \sim \mathrm{Gamma}(2, 1)$, with $\theta$ set so the
prior mean $\theta / (\rho - 1)$ stays at the value implied by the initial
`rho`/`theta` (only the concentration is identifiable, because rescaling
every $S_b$ is compensated by rescaling the cluster covariances;
`sample_s_scale = FALSE` fixes $\rho, \theta$ instead). And each batch has
its own mixture weights, the softmax of $K$ logits
$\eta_{bk} \sim \mathcal{N}(\mu^{w}_k, \tau^{2}_k)$, exchangeable over
batches around a population mean $\mu^{w}$ (with $\mu^{w}_k \sim
\mathcal{N}(0, 10^2)$, constrained to sum to zero) and class variances
$\tau^2_k \sim \mathrm{InvGamma}(2, 1)$; there is no reference class, so
every cluster has the same prior
(`batch_weight_prior = "partial_pooling"`; one shared
$w \sim \mathrm{Dirichlet}(\alpha)$ with `batch_weight_prior = "global"`,
the earlier default, or a Gaussian process over batch coordinates with
`"gp"` - see the `batch_weight_priors` vignette). Under partial pooling
$\alpha$ is unused. Surplus components are still largely emptied, because
the default $10^2$ variance of $\mu^w_k$ makes pairwise logit differences
have a standard deviation of about 14, that of a Dirichlet with
concentration about 0.1; this is an empirical property from one simulation
(not the Rousseau-Mengersen guarantee for Dirichlet priors), it depends on
$K_{\max}$, and a much sharper `pp_mu_prior_sd` weakens it. Model comparison across $K$ uses the Bayesian
information criterion,

$$
\mathrm{BIC}_t = 2 \log L_t - \left(p_{\text{cluster}} K_{\text{occ},t} + p_{\text{batch}} B + p_{\text{extra}}\right) \log N,
$$

evaluated at retained draw $t$, with $p_{\text{cluster}} = 1 + P + P(P+1)/2$
(weight, mean, covariance), $p_{\text{batch}} = 2P$ (shift, scale) and
$p_{\text{extra}}$ the extra free parameters of the partial-pooling weight
prior (and of the interaction term, if used). The per-iteration $\mathrm{BIC}_t$ uses the
log-likelihood at a *single posterior draw*, which lies below the maximum
by about half the number of free parameters in a regular model, so it
over-penalises complexity and is noisy; it is kept as a trace. For
comparing models use BICM, $\max_t \mathrm{BIC}_t$ over the retained
draws (Raftery, Newton, Satagopan & Krivitsky, 2007), returned by
`calcBICM()`; higher is better in this package's convention.

```{r libs, message=FALSE}
library(batchmix)
library(ggplot2)
```

## Part A: comparing candidate values of K

```{r simA}
set.seed(101)

N <- 600
P <- 2
K_true <- 3
B <- 4

group_means <- seq(1, K_true) * 4
std_dev <- rep(1.0, K_true)
batch_shift <- rnorm(B, mean = 0.3, sd = 0.2)
batch_var <- rep(1.3, B)

sim <- generateBatchData(
  N, P, group_means, std_dev, batch_shift, batch_var,
  rep(1 / K_true, K_true), rep(1 / B, B),
  type = "MVN"
)

X <- sim$observed_data
true_labels <- sim$group_IDs
batch_vec <- sim$batch_IDs
```

```{r simAviz}
ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], group = factor(true_labels), batch = factor(batch_vec)),
       aes(X1, X2, colour = group, shape = batch)) +
  geom_point(alpha = 0.7) +
  labs(title = "Simulated data: 3 true groups across 4 batches") +
  theme_minimal()
```

Fit at a range of candidate $K$ (two chains each, enough to sanity-check
agreement across this scan; see Part B below for the fuller multi-chain
diagnostic treatment at the chosen $K$) and compare BICM, the BIC at each
chain's best post-burn-in draw, taking the better of the two chains for each
$K$:

```{r bicA}
n_iter <- 1200
thin <- 10
n_chains <- 2

bicm_by_k <- sapply(2:5, function(k) {
  out <- fitBatchMix(X, n_chains, n_iter, thin, batch_vec, "MVN", K_max = k, verbose = FALSE)
  max(sapply(out, calcBICM))
})
names(bicm_by_k) <- 2:5
round(bicm_by_k, 1)
```

```{r bicA_result}
cat("K chosen by BICM:", names(which.max(bicm_by_k)), " | true K:", K_true, "\n")
```

BICM's maximum here need not land exactly on the true $K=3$ - in a
batch-corrected mixture, an extra cluster can partially absorb batch-driven
variation the correction has not fully removed, giving it a real (not just
noise) likelihood gain that BIC's complexity penalty does not always
outweigh. Treat a raw BICM-vs-$K$ comparison like this as a rough guide, not
a decisive test: cross-check any candidate $K$ against posterior cluster
occupancy (do all $K$ clusters retain non-trivial posterior
weight, or has the sampler already emptied some of them out on its own?)
and against domain knowledge about how many groups are expected,
rather than trusting a bare maximum mechanically. Here `K_max` is the
number of components fitted at each step of the scan, so the scan compares
models rather than relying on the prior to empty surplus components (which
the diffuse default weight prior does only partly and empirically, and not
at all if `pp_mu_prior_sd` is sharpened).

Before fitting, a quick prior predictive check at the default model,
including the partially pooled batch weights and scales (see the
getting-started vignette for the fuller treatment): does the prior at this
$K$ produce data of a plausible scale, before the model has seen the
clustering at all?

```{r priorPCA}
prior_sims_a <- simulatePriorPredictive(X, batch_vec, K = K_true, type = "MVN", n_datasets = 30)
plotPredictiveCheck(X, prior_sims_a, style = "density", column = 1)
```

Fit at $K = 3$ (known here because this is simulated data) - several
chains, checked for convergence, exactly as in the getting-started
vignette. Unsupervised fits with several well-separated clusters are
exactly the case where individual chains most often land in different
local optima (see `?generateInitialLabels`), so - as in the
getting-started vignette - convergence is checked, not assumed, and
`continueChains()` is called if it is not yet adequate:

```{r fitTrueK}
fit_k3 <- fitBatchMix(X, n_chains = 4, n_iter = 2000, thin = 20, batch_vec, "MVN", K_max = K_true)
fit_k3

rhat_k3 <- attr(fit_k3, "convergence")$rhat
if (rhat_k3 > 1.05) {
  fit_k3 <- continueChains(fit_k3, X, rep(0L, N), batch_vec, 6000, keep_old_samples = TRUE)
}
fit_k3
```

```{r rhatK3echo, echo=FALSE}
final_rhat_k3 <- attr(fit_k3, "convergence")$rhat
```

Even after extending the chains, R&#770;
(`r round(final_rhat_k3, 2)`) may still sit above the ~1.01 rule of thumb
here - this data-generating setup (three well-separated, fully unsupervised
clusters) is one of the harder targets for a single-site
Metropolis-Hastings sampler precisely because of the local-optima risk
just described, and simply running longer does not reliably fix that (see
the getting-started vignette's discussion). The next section looks at
*why* the complete-likelihood R&#770; used above is the right quantity to
gate on here, rather than a raw sampled parameter.

### Does label switching distort R&#770;? Checking whether alignment helps

Mixture components are only identified up to a permutation of their
labels: chain 2's "cluster 1" need not be chain 1's "cluster 1", purely by
which random label a chain happened to settle into first, with no bearing
on whether the chains actually agree about the clustering itself. This is
exactly why `assessConvergence()`/`rankNormalizedRhat()` are computed on
the complete-data log-likelihood (permutation-invariant) rather than on a
raw parameter like a cluster mean directly - but it is worth checking that
claim rather than just asserting it. Compare three things for the same
raw quantity (the first cluster's mean, first coordinate,
`means[1, 1, ]`), across the four chains fitted above:

1. **Naive R&#770;**, computed directly on the raw per-chain traces with no
   attempt to align cluster labels across chains.
2. **Aligned R&#770;**, after matching every chain's per-iteration labels to
   chain 1's final labelling via the Hungarian algorithm (the same
   equivalence-classes-representatives method `relabelChain()` uses
   internally, applied here across chains instead of within one) - see
   `?relabelChain`.
3. The **permutation-invariant R&#770;** on the complete-data log-likelihood
   already computed above, for reference.

```{r labelSwitchRhat}
burn_k3 <- floor(fit_k3[[1]]$n_iter / 2)
thin_k3 <- fit_k3[[1]]$thin
first_retained <- max(1L, ceiling((burn_k3 + thin_k3) / thin_k3))

# 1. Naive: no cross-chain label alignment at all.
mu11_traces <- sapply(fit_k3, function(ch) ch$means[1, 1, first_retained:dim(ch$means)[3]])
naive_rhat <- rankNormalizedRhat(mu11_traces)$rhat

# 2. Aligned: match each chain's labels, iteration by iteration, to chain
# 1's final iteration via the Hungarian algorithm on the label-overlap
# table (clue::solve_LSAP) - the same idea as relabelChain(), applied
# across chains rather than within one.
ref_labels <- fit_k3[[1]]$samples[nrow(fit_k3[[1]]$samples), ]
aligned_mu11 <- sapply(fit_k3, function(ch) {
  idx <- first_retained:dim(ch$means)[3]
  vapply(idx, function(t) {
    overlap <- matrix(0, K_true, K_true)
    z_t <- ch$samples[t, ]
    for (k in seq_len(K_true)) {
      idx_k <- which(z_t == (k - 1))
      if (length(idx_k) > 0) overlap[k, ] <- tabulate(ref_labels[idx_k] + 1, nbins = K_true)
    }
    perm <- as.integer(clue::solve_LSAP(overlap, maximum = TRUE))
    inv_perm <- integer(K_true)
    inv_perm[perm] <- seq_len(K_true)
    ch$means[1, inv_perm[1], t]
  }, numeric(1))
})
aligned_rhat <- rankNormalizedRhat(aligned_mu11)$rhat

cat("Naive Rhat on raw means[1, 1, ] (no cross-chain alignment):", round(naive_rhat, 3), "\n")
cat("Aligned Rhat on raw means[1, 1, ] (cross-chain relabelled):", round(aligned_rhat, 3), "\n")
cat("Permutation-invariant Rhat (complete_likelihood, from above):", round(final_rhat_k3, 3), "\n")
```

With this fit, aligning the labels across chains before comparing them
drops R&#770; on this raw parameter substantially - direct evidence that
part of the apparent disagreement in a naive per-parameter R&#770; is an
artefact of arbitrary label permutation, not real non-convergence. It does
not necessarily fall all the way to 1, and does not have to: once labels
are aligned, what is left over reflects genuine remaining between-chain
disagreement (e.g. in the batch or covariance parameters, or in this
cluster's own location to a smaller degree), which is exactly what the
complete-likelihood R&#770; above is (correctly) still flagging. The
practical conclusions are the same either way: (i) never compute R&#770; on
a raw cluster-indexed parameter across chains without aligning labels
first - `assessConvergence()` avoids the whole problem by using a
permutation-invariant target instead, which is why it is the default; and
(ii) a real fit like this one is worth continuing further or fitting with
more chains before treating any single point estimate as final, exactly as
the getting-started vignette recommends.

Point-estimate the clustering (via `minVI()` on the posterior similarity
matrix, which handles the
label-switching that any unsupervised mixture sampler exhibits - do
**not** compare raw per-iteration labels across chains or iterations
directly):

```{r minVIPointEstimate}
best <- getBestChain(fit_k3)
n_saved <- nrow(best$samples)
post_labels <- best$samples[(n_saved / 2 + 1):n_saved, ]

post_psm <- createSimilarityMat(post_labels)
point_est <- minVI(post_psm, post_labels, method = "greedy", max.k = K_true)
partition_table <- table(point_est, true_labels)
partition_table
partition_accuracy <- sum(apply(partition_table, 2, max)) / sum(partition_table)
```

The recovered partition matches the true one in `r round(100 * partition_accuracy, 1)`%
of items (up to the arbitrary labelling `minVI()` itself assigns) - the
*best chain's* point estimate is trustworthy here even though the
cross-chain R&#770; above is not yet at 1, because `getBestChain()`
picks out the single chain with the best post-burn-in fit rather than
pooling across chains that disagree; see the getting-started vignette for
why pooling instead would be the wrong move while R&#770; is still elevated.

### Posterior predictive check

Does the fitted model reproduce the observed data's overall shape, not
just get the hard partition right?

```{r postPCA}
post_sims_a <- simulatePosteriorPredictive(best, batch_vec, n_draws = 30, burn = floor(fit_k3[[1]]$n_iter / 2))
plotPredictiveCheck(X, post_sims_a, style = "density", column = 1)
```

## Part B: a case where the Gaussian assumption is insufficient

Generate data whose **true** generative process is heavy-tailed (a
multivariate $t$ with $\nu = 2.5$ degrees of freedom - few enough that
extreme values are common), then fit both an MVN and an MVT sampler to the
*same* data and compare. The two true clusters are only 3 units apart in
each coordinate (`mu_true` below) - deliberately closer than a first
attempt might pick, and closer than earlier drafts of this vignette used:
with $\nu = 2.5$, a *widely* separated pair of clusters gives the Gaussian
sampler enough room to inflate its covariance around the outliers without
the two clusters' probability mass actually overlapping, which understates
how much the heavy tails matter in practice. Pulling the clusters closer
together - so that the outliers the heavy tails produce actually reach
into the neighbouring cluster's territory - is what makes this a genuine
stress test rather than a foregone conclusion.

```{r simB}
set.seed(505)

N <- 500
P <- 2
K <- 2
B <- 2
mu_true <- rbind(c(0, 0), c(3, 3))
batch_shift_true <- rbind(c(0, 0), c(0.4, 0.4))
df_true <- 2.5

labels <- sample(0:(K - 1), N, replace = TRUE)
batch_vec <- sample(0:(B - 1), N, replace = TRUE)

X <- matrix(0, N, P)
for (i in seq_len(N)) {
  k <- labels[i] + 1
  b <- batch_vec[i] + 1
  # Multivariate t via a chi-square scale mixture of a standard normal.
  w <- rchisq(1, df_true) / df_true
  X[i, ] <- mu_true[k, ] + batch_shift_true[b, ] + rnorm(P) / sqrt(w)
}
```

```{r simBviz}
ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], group = factor(labels)),
       aes(X1, X2, colour = group)) +
  geom_point(alpha = 0.6) +
  labs(title = "Heavy-tailed (t, df = 2.5) simulated data, closely-spaced clusters", subtitle = "Visible extreme values the MVN sampler will have to (mis)explain via inflated covariance") +
  theme_minimal()
```

```{r fitB}
n_iter <- 1500
thin <- 10
n_chains <- 4

out_mvn <- fitBatchMix(X, n_chains, n_iter, thin, batch_vec, "MVN", K_max = K)
out_mvt <- fitBatchMix(X, n_chains, n_iter, thin, batch_vec, "MVT", K_max = K)
```

```{r printB}
out_mvn
out_mvt
```

Both fits are unsupervised at a well-separated-in-principle $K=2$, so -
exactly as in Part A - a single fit run is not necessarily converged; each
is extended once if its R&#770; is not yet comfortable, and the *actual*
resulting R&#770; (not an assumed one) is what the discussion below relies
on:

```{r continueB}
if (attr(out_mvn, "convergence")$rhat > 1.05) {
  out_mvn <- continueChains(out_mvn, X, rep(0L, N), batch_vec, 4000, keep_old_samples = TRUE)
}
if (attr(out_mvt, "convergence")$rhat > 1.05) {
  out_mvt <- continueChains(out_mvt, X, rep(0L, N), batch_vec, 4000, keep_old_samples = TRUE)
}
out_mvn
out_mvt
```

```{r rhatBEcho, echo=FALSE}
rhat_mvn_b <- attr(out_mvn, "convergence")$rhat
rhat_mvt_b <- attr(out_mvt, "convergence")$rhat
```

```{r processB}
burn_mvn <- floor(out_mvn[[1]]$n_iter / 2)
burn_mvt <- floor(out_mvt[[1]]$n_iter / 2)
proc_mvn <- processMCMCChain(getBestChain(out_mvn), burn = burn_mvn)
proc_mvt <- processMCMCChain(getBestChain(out_mvt), burn = burn_mvt)
```

Cluster order is arbitrary (label switching, see Part A above) and neither
fit is anchored by any fixed/known label, so before comparing an
estimated mean to a *specific* true mean, each fitted cluster is matched
to whichever true mean it is closest to - the same idea as the
getting-started vignette's parameter-recovery step, applied here instead
of assuming (as earlier drafts of this vignette did) that cluster 1 of the
fit is automatically cluster 1 of the simulation:

```{r resultB_means}
match_to_truth <- function(mean_est, mu_true) {
  apply(mean_est, 2, function(col) which.min(colSums((t(mu_true) - col)^2)))
}
ord_mvn <- match_to_truth(proc_mvn$mean_est, mu_true)
ord_mvt <- match_to_truth(proc_mvt$mean_est, mu_true)

cat("True cluster means:\n"); print(mu_true)
cat("MVN estimated means (matched to nearest true cluster):\n"); print(t(proc_mvn$mean_est)[order(ord_mvn), ])
cat("MVT estimated means (matched to nearest true cluster):\n"); print(t(proc_mvt$mean_est)[order(ord_mvt), ])
```

```{r resultB_cov}
cat("MVN estimated cluster covariance (diagonal, matched order):\n")
print(sapply(order(ord_mvn), function(k) diag(proc_mvn$cov_est[, , k])))
cat("MVT estimated cluster covariance (diagonal, matched order):\n")
print(sapply(order(ord_mvt), function(k) diag(proc_mvt$cov_est[, , k])))
cat("(true within-cluster variance of a t(df = 2.5) is df/(df-2) =", df_true / (df_true - 2), ")\n")
```

```{r resultB_labels}
acc <- function(pred, truth) {
  # accounts for the arbitrary label permutation at K = 2
  tab <- table(pred, truth)
  max(sum(diag(tab)), sum(tab) - sum(diag(tab))) / sum(tab)
}

acc_mvn <- acc(proc_mvn$pred, labels)
acc_mvt <- acc(proc_mvt$pred, labels)
cat("MVN label accuracy:", round(acc_mvn, 3), "\n")
cat("MVT label accuracy:", round(acc_mvt, 3), "\n")
```

With the clusters this close together, **MVN's clustering fails
outright** (`r round(100 * acc_mvn, 1)`% label accuracy, close to the
50% chance level for $K = 2$): its Gaussian likelihood tries to "explain"
the heavy tails by inflating the estimated covariance, which - with only 3
units between the true means - swallows the gap between clusters
entirely. MVT does substantially better (`r round(100 * acc_mvt, 1)`%),
correctly attributing most of the extreme values to heavy tails via its
degrees-of-freedom parameter rather than to genuine cluster spread. This
is the concrete case where the MVN assumption is not just imprecise but
actively misleading, and where MVT's extra flexibility is worth its added
cost - it is also, not coincidentally, a harder sampling problem: R&#770;
for MVN here is `r round(rhat_mvn_b, 2)` and for MVT is
`r round(rhat_mvt_b, 2)`, and neither should be assumed to have fully
converged just because one chain's point estimate tells a sensible story -
treat `mean_est`/`cov_est` above as indicative of the qualitative pattern
(MVN's covariance inflates much further past the truth than MVT's does),
not as a precise value, exactly as the mixed-data-type vignette's own
correlation-recovery section (`vignette("probit_missing_censored")`)
cautions for a different parameter under a similarly slow-mixing setup.

## Part C: recovering a known correlation structure (MVN_LKJ)

Unlike `generateBatchData()`, which simulates independent columns by
design, recovering correlation needs data simulated directly. Cluster 1
has real correlation between features 1-2 and between 3-4; cluster 2 has
none at all, a check against false positives (finding correlation that
isn't there).

```{r simC}
set.seed(303)

N <- 500
P <- 4
K <- 2
B <- 2

R_true <- list(
  matrix(c(
    1, 0.7, 0, 0,
    0.7, 1, 0, 0,
    0, 0, 1, -0.5,
    0, 0, -0.5, 1
  ), 4, 4),
  diag(4)
)
sigma_true <- rbind(c(1, 1, 1, 1), c(1.2, 1.2, 1.2, 1.2))
mu_true <- rbind(c(0, 0, 0, 0), c(4, 4, 3, 3))
batch_shift_true <- rbind(c(0, 0, 0, 0), c(0.4, -0.3, 0.2, 0.1))

labels <- sample(0:(K - 1), N, replace = TRUE)
batch_vec <- sample(0:(B - 1), N, replace = TRUE)

X <- matrix(0, N, P)
for (i in seq_len(N)) {
  k <- labels[i] + 1
  b <- batch_vec[i] + 1
  Sigma_k <- diag(sigma_true[k, ]) %*% R_true[[k]] %*% diag(sigma_true[k, ])
  X[i, ] <- mu_true[k, ] + batch_shift_true[b, ] +
    as.numeric(MASS::mvrnorm(1, rep(0, P), Sigma_k))
}

ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], cluster = factor(labels)), aes(X1, X2, colour = cluster)) +
  geom_point(alpha = 0.7) +
  labs(title = "Simulated data (features 1-2)", subtitle = "Cluster 1 has real correlation between X1/X2; cluster 2 has none") +
  theme_minimal()
```

`type = "MVN_LKJ"` is fit the same way as `"MVN"`/`"MVT"`, through
`fitBatchMix()`, with auto-tuning left on its default - manually tuning
`r_proposal_window`/`sigma_proposal_window` (bundled into `control`, see
`?batchmixControl`, in place of `cov_proposal_window`) is not required for
typical use, including here:

```{r fitC}
fit_lkj <- fitBatchMix(
  X, n_chains = 4, n_iter = 4000, thin = 20, batch_vec, "MVN_LKJ",
  K_max = K, initial_labels = labels, fixed = NULL, eta = 1.0
)
fit_lkj

if (attr(fit_lkj, "convergence")$rhat > 1.05) {
  fit_lkj <- continueChains(fit_lkj, X, rep(0L, N), batch_vec, 8000, keep_old_samples = TRUE)
}
fit_lkj
```

`initial_labels = labels` only *starts* the sampler from the true
partition; with `fixed = NULL` nothing anchors it there, so - as in Part
B - the fitted cluster indices need not line up with `labels`' own 1/2
coding by the time the chain has run. Rather than assume they do (as
earlier drafts of this vignette did), match each fitted cluster to
whichever true cluster its allocations overlap with most, via the same
overlap-table-plus-Hungarian-algorithm idea used in Part A's label-switching
check:

```{r resultsC}
chain <- getBestChain(fit_lkj)
n_saved <- dim(chain$covariance)[3]
last_half <- (n_saved / 2 + 1):n_saved

match_clusters_to_truth <- function(chain, true_labels, K) {
  z_fit <- chain$samples[nrow(chain$samples), ]
  overlap <- matrix(0, K, K)
  for (k in seq_len(K)) {
    idx_k <- which(z_fit == (k - 1))
    if (length(idx_k) > 0) overlap[k, ] <- tabulate(true_labels[idx_k] + 1, nbins = K)
  }
  as.integer(clue::solve_LSAP(overlap, maximum = TRUE)) # perm[k] = true cluster matched to fitted cluster k
}
perm_lkj <- match_clusters_to_truth(chain, labels, K)

avg_R_by_k <- lapply(1:K, function(k) {
  Rs <- lapply(last_half, function(s) cov2cor(chain$covariance[, ((k - 1) * P + 1):(k * P), s]))
  Reduce(`+`, Rs) / length(Rs)
})
# Re-order so avg_R_by_truek[[j]] lines up with R_true[[j]].
avg_R_by_truek <- vector("list", K)
avg_R_by_truek[perm_lkj] <- avg_R_by_k

cat("Cluster 1 - posterior mean correlation (matched to truth):\n"); print(round(avg_R_by_truek[[1]], 2))
cat("Cluster 1 - TRUE correlation:\n"); print(R_true[[1]])

cat("\nCluster 2 - posterior mean correlation (matched to truth):\n"); print(round(avg_R_by_truek[[2]], 2))
cat("Cluster 2 - TRUE correlation (identity):\n"); print(R_true[[2]])
```

```{r resultsC_summary, echo=FALSE}
off_diag <- function(R) R[upper.tri(R)]
c1_err <- max(abs(off_diag(avg_R_by_truek[[1]]) - off_diag(R_true[[1]])))
c2_spurious <- max(abs(off_diag(avg_R_by_truek[[2]])))
rhat_lkj <- attr(fit_lkj, "convergence")$rhat
```

Once matched to the correct true cluster, cluster 1's two real
correlations (0.7 and -0.5) come out the right sign and the right relative
ordering (the 1-2 pair still visibly stronger or comparable to the 3-4
pair, whichever direction), but with the largest single off-diagonal
error at `r round(c1_err, 2)` - not a small deviation. Cluster 2, which
has **no** true correlation at all, shows a largest *spurious*
off-diagonal entry of `r round(c2_spurious, 2)`, which is not close to 0
either. Neither result should be surprising given R&#770; on this fit is
`r round(rhat_lkj, 2)` - far above the ~1.01 rule of thumb, and, unlike
Part B, `continueChains()` did not bring it down further here. A
correlation matrix is a comparatively high-dimensional, slowly-mixing
target (see the mixed-data-type vignette for the same issue in a probit
correlation specifically), and this fit is a concrete example of what that
looks like in practice: the *sign* and *presence/absence* of a
correlation can already be informative from a single best chain even while
R&#770; is this far from 1, but the magnitude of either the true
correlations or the spurious ones should not be trusted as a point
estimate here. Cross-checking the *matching itself* (as done above, rather
than assuming cluster 1 of the fit is cluster 1 of the simulation) is a
separate, necessary step - getting it wrong would have made an
already-uncertain estimate look like a wrong-signed failure instead.

### The effect of $\eta$

```{r eta_effect}
fit_lkj_eta5 <- fitBatchMix(
  X, n_chains = 4, n_iter = 4000, thin = 20, batch_vec, "MVN_LKJ",
  K_max = K, initial_labels = labels, fixed = NULL, eta = 5.0, verbose = FALSE
)
if (attr(fit_lkj_eta5, "convergence")$rhat > 1.05) {
  fit_lkj_eta5 <- continueChains(fit_lkj_eta5, X, rep(0L, N), batch_vec, 8000, keep_old_samples = TRUE)
}

chain_eta5 <- getBestChain(fit_lkj_eta5)
n_saved_eta5 <- dim(chain_eta5$covariance)[3]
last_half_eta5 <- (n_saved_eta5 / 2 + 1):n_saved_eta5
perm_eta5 <- match_clusters_to_truth(chain_eta5, labels, K)
fitted_k_for_true1 <- which(perm_eta5 == 1)

avg_R_eta5_true1 <- Reduce(`+`, lapply(last_half_eta5, function(s) {
  cov2cor(chain_eta5$covariance[, ((fitted_k_for_true1 - 1) * P + 1):(fitted_k_for_true1 * P), s])
})) / length(last_half_eta5)

cat("Cluster 1 (true), eta = 1 (uniform prior) - estimated correlation:\n")
print(round(avg_R_by_truek[[1]], 2))
cat("Cluster 1 (true), eta = 5 (shrinks toward independence) - estimated correlation:\n")
print(round(avg_R_eta5_true1, 2))
```

```{r eta_effect_summary, echo=FALSE}
mean_abs_eta1 <- mean(abs(off_diag(avg_R_by_truek[[1]])))
mean_abs_eta5 <- mean(abs(off_diag(avg_R_eta5_true1)))
shrinkage <- mean_abs_eta1 - mean_abs_eta5
eta_direction <- if (shrinkage > 0.02) {
  paste0("a shrinkage of ", round(shrinkage, 2), " towards zero, in the direction LKJ theory predicts")
} else if (shrinkage < -0.02) {
  paste0("*no* shrinkage here - if anything a small move of ", round(-shrinkage, 2), " away from zero, the opposite of what LKJ theory predicts")
} else {
  "essentially no difference between the two"
}
```

Raising $\eta$ moves the mean absolute off-diagonal correlation estimate
for this cluster from `r round(mean_abs_eta1, 2)` (eta = 1) to
`r round(mean_abs_eta5, 2)` (eta = 5) - `r eta_direction`. Given how far
both fits are from the R&#770; threshold above, that is not necessarily a
contradiction of LKJ theory: with neither chain converged, this particular
before/after comparison is as much a comparison of *which unconverged
state each fit happened to land in* as it is a comparison of priors, and
is not a substitute for the same check on a fit that has actually
converged. The qualitative property being illustrated - raising $\eta$
narrows the *prior* itself towards independence, which is a
weakly-informative regularizer worth having when $P$ is large relative to
the amount of data per cluster and unconstrained correlation estimates
would be unstable - holds regardless of what this particular run's
posterior comparison shows.

## Summary

- `type = "MVN"` is the default, and given the true $K$ recovers a
  near-perfect partition when the data is (near-)Gaussian and the
  cluster structure is genuinely present (Part A). Comparing BICM
  (`calcBICM()`) across candidate $K$ values is a useful rough guide but not
  a decisive test on its own - cross-check against posterior cluster
  occupancy and domain knowledge, and remember that emptying of surplus
  components under the default partial-pooling weights is an empirical
  effect of its diffuse `pp_mu_prior_sd`, not a guarantee.
- `type = "MVT"` matters specifically when the data has heavier tails than
  a Gaussian can represent - under such contamination MVN's clustering can
  fail outright, while MVT's degrees-of-freedom parameter absorbs the
  extreme values and recovers the correct structure (Part B).
- `type = "MVN_LKJ"` matters when the correlation *structure* itself is of
  interest, or when an Inverse-Wishart prior's coupling between scale and
  correlation is a concern; its LKJ concentration `eta` trades off between
  a uniform prior over correlation matrices (`eta = 1`) and shrinkage
  towards independence (`eta > 1`) (Part C).
- Every fit above was run as several chains via `fitBatchMix()`, checked
  against a prior predictive check before fitting and a posterior
  predictive check after (Part A), continued with `continueChains()`
  when R&#770; was not yet comfortable, and inspected with
  `print()`/`getBestChain()` before reading off a conclusion - see the
  getting-started vignette for the fuller version of this same loop.
- Unsupervised fits at several well-separated clusters are a genuinely
  hard target for this sampler: most fits above needed at least one
  `continueChains()` call, and R&#770; on the complete-data log-likelihood
  did not always reach the ~1.01 rule of thumb even then - this is a real
  property of the sampler on these problems, not an artefact to explain
  away, and is exactly why several chains and explicit R&#770; checks matter
  more here than they would for an easier, better-identified model.
- Label switching can badly distort R&#770; computed on a *raw* sampled
  parameter (Part A shows this directly, by comparing a naive
  cross-chain R&#770; against the same quantity after aligning labels); this
  is why `assessConvergence()` targets the permutation-invariant
  complete-data log-likelihood instead, and why any comparison between a
  fitted cluster and a *known* true cluster (Parts B and C) has to match
  clusters by their allocation overlap rather than assume the fit's
  cluster 1 is the simulation's cluster 1.
