---
title: "Choosing a cluster covariance model: robustness and correlation"
author: "batchmix"
output: rmarkdown::html_vignette
description: |
  Comparing the three continuous-data covariance models batchmix offers -
  MVN, MVT and MVN_LKJ - against data with a known ground truth: comparing
  candidate cluster counts via BIC, does MVT's robustness matter for
  heavy-tailed data, and does the LKJ prior recover a known correlation
  structure.
vignette: >
  %\VignetteIndexEntry{Choosing a cluster covariance model: robustness and correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
```

## Purpose

batchmix offers three continuous-data cluster covariance models, sharing
the same batch-correction machinery and directly comparable as fits to the
same data:

- `type = "MVN"`: multivariate normal, Inverse-Wishart covariance prior.
  The default, and the right choice unless one of the two reasons below
  applies.
- `type = "MVT"`: multivariate $t$, with a per-cluster degrees-of-freedom
  parameter learned from the data. Robust to heavy-tailed/outlier-
  contaminated data, where MVN's Gaussian likelihood is not just imprecise
  but actively misleading (Part B below).
- `type = "MVN_LKJ"`: multivariate normal with an LKJ prior on the cluster
  *correlation* matrix and independent log-normal priors on the marginal
  *scales*, decoupling two things an Inverse-Wishart prior couples
  together (Part C below).

This vignette validates each against data with a **known ground truth**
from simulation - for checking whether a sampler recovers a *known*
generative model, this is the strictest possible test: the alternative (a
real dataset with expert-assigned labels) has its own label noise, so a
sampler could look "wrong" for reasons that have nothing to do with the
implementation.

## The hierarchical model

For item $i = 1, \dots, N$, with cluster label $c_i \in \{1,\dots,K\}$ and
batch label $b_i \in \{1,\dots,B\}$:

$$
x_i \mid c_i = k,\ b_i = b \ \sim\
\begin{cases}
\mathcal{N}\!\left(\mu_k + m_b,\ S_b \bullet \Sigma_k\right) & \texttt{"MVN"} \\[4pt]
t_{\nu_k}\!\left(\mu_k + m_b,\ S_b \bullet \Sigma_k\right) & \texttt{"MVT"} \\[4pt]
\mathcal{N}\!\left(\mu_k + m_b,\ S_b \bullet (\Delta_k R_k \Delta_k)\right) & \texttt{"MVN\_LKJ"}
\end{cases}
$$

where $S_b = \mathrm{diag}(S_{b,1}, \dots, S_{b,P})$ is a per-batch,
per-feature multiplicative scale, $m_b$ is a per-batch additive shift, and
$\bullet$ rescales only the *diagonal* of the cluster covariance,
$(S_b \bullet \Sigma_k)_{pp} = S_{b,p}\,\Sigma_{k,pp}$ and
$(S_b \bullet \Sigma_k)_{pq} = \Sigma_{k,pq}$ for $p \neq q$ - i.e. the batch
scale inflates each feature's marginal variance but leaves the cluster's raw
covariances between features untouched (this is deliberate: batch effects
are modelled as per-feature measurement noise, not a change in how features
correlate). This is *not* the same as the full congruence transform
$S_b^{1/2} \Sigma_k S_b^{1/2}$ except in the special case where $\Sigma_k$ is
itself diagonal (no correlation between features): whenever $\Sigma_k$ has
off-diagonal entries, only the diagonal is rescaled, so the batch-corrected
*correlation* between two features still shifts with $S_b$ (their raw
covariance is unchanged while their variances grow), even though the raw
covariance entry itself does not move. For `"MVN"`/`"MVT"`, $\mu_k, \Sigma_k$
are the cluster mean and covariance with $\Sigma_k \sim \mathrm{IW}(\nu,
\Psi)$. `"MVN_LKJ"` decomposes the
covariance as $\Delta_k = \mathrm{diag}(\sigma_{k,1}, \dots, \sigma_{k,P})$
with **independent** priors on the correlation matrix and the marginal
scales,

$$
R_k \sim \mathrm{LKJ}(\eta), \qquad \log \sigma_{k,p} \sim \mathcal{N}(\beta, \xi^2),
$$

which lets a weakly-informative guess at the scales be stated without
implying anything about the correlations, or vice versa - something an
Inverse-Wishart prior makes hard, since it couples the two (Barnard,
McCulloch & Meng, 2000, *Statistica Sinica*; Tokuda, Goodrich, Van
Mechelen, Gelman & Meng, 2011, "Visualizing Distributions of Covariance
Matrices"). $\mathrm{LKJ}(\eta)$ (Lewandowski, Kurowicka & Joe, 2009, *J.
Multivariate Analysis*) has density $p(R \mid \eta) \propto \det(R)^{\eta -
1}$: $\eta = 1$ is uniform over the space of valid correlation matrices,
$\eta > 1$ concentrates mass towards the identity (shrinks correlations
towards 0) - it is the standard modern choice for this decomposition (Stan
and brms use it the same way).

The remaining priors are shared by every type:

$$
\mu_k \mid \Sigma_k \sim \mathcal{N}\!\left(\xi, \Sigma_k / \kappa\right), \qquad
m_b \sim \mathcal{N}(0, \delta^2 \lambda^2), \qquad
S_{b,p} - 1 \sim \mathrm{InvGamma}(\rho, \theta), \qquad
w \sim \mathrm{Dirichlet}(\alpha),
$$

with $\nu_k - \nu_{\min} \sim \mathrm{Gamma}(\psi, \chi)$ additionally for
`"MVT"`. Model comparison across $K$ uses the Bayesian information
criterion,

$$
\mathrm{BIC} = 2 \log \hat{L} - \left(p_{\text{cluster}} K_{\text{occ}} + p_{\text{batch}} B\right) \log N,
$$

with $p_{\text{cluster}} = 1 + P + P(P+1)/2$ (weight, mean, covariance) and
$p_{\text{batch}} = 2P$ (shift, scale); higher is better in this package's
convention.

```{r libs, message=FALSE}
library(batchmix)
library(salso)
library(ggplot2)
```

## Part A: comparing candidate values of K

```{r simA}
set.seed(101)

N <- 600
P <- 2
K_true <- 3
B <- 4

group_means <- seq(1, K_true) * 4
std_dev <- rep(1.0, K_true)
batch_shift <- rnorm(B, mean = 0.3, sd = 0.2)
batch_var <- rep(1.3, B)

sim <- generateBatchData(
  N, P, group_means, std_dev, batch_shift, batch_var,
  rep(1 / K_true, K_true), rep(1 / B, B),
  type = "MVN"
)

X <- sim$observed_data
true_labels <- sim$group_IDs
batch_vec <- sim$batch_IDs
```

```{r simAviz}
ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], group = factor(true_labels), batch = factor(batch_vec)),
       aes(X1, X2, colour = group, shape = batch)) +
  geom_point(alpha = 0.7) +
  labs(title = "Simulated data: 3 true groups across 4 batches") +
  theme_minimal()
```

Fit at a range of candidate $K$ (two chains each, enough to sanity-check
agreement across this scan; see Part B below for the fuller multi-chain
diagnostic treatment at the chosen $K$) and compare mean post-burn-in BIC:

```{r bicA}
n_iter <- 1200
thin <- 10
n_chains <- 2

bic_by_k <- sapply(2:5, function(k) {
  out <- fitBatchMix(X, n_chains, n_iter, thin, batch_vec, "MVN", K_max = k, verbose = FALSE)
  mean(sapply(out, function(o) {
    n <- length(o$BIC)
    mean(o$BIC[(n / 2):n])
  }))
})
names(bic_by_k) <- 2:5
round(bic_by_k, 1)
```

```{r bicA_result}
cat("K chosen by BIC:", names(which.max(bic_by_k)), " | true K:", K_true, "\n")
```

BIC's maximum here need not land exactly on the true $K=3$ - in a
batch-corrected mixture, an extra cluster can partially absorb batch-driven
variation the correction has not fully removed, giving it a real (not just
noise) likelihood gain that BIC's complexity penalty does not always
outweigh. Treat a raw BIC-vs-$K$ comparison like this as a rough guide, not
a decisive test: cross-check any candidate $K$ against posterior cluster
occupancy (do all $K$ clusters retain non-trivial posterior
weight, or has the sampler already emptied some of them out on its own?)
and against domain knowledge about how many groups are expected,
rather than trusting a bare maximum mechanically.

Fit at $K = 3$ (known here because this is simulated data) - several
chains, checked for convergence, exactly as in the getting-started
vignette:

```{r fitTrueK}
fit_k3 <- fitBatchMix(X, n_chains = 4, n_iter = 2000, thin = 20, batch_vec, "MVN", K_max = K_true)
fit_k3
```

Point-estimate the clustering (via `salso`, which handles the
label-switching that any unsupervised mixture sampler exhibits - do
**not** compare raw per-iteration labels across chains or iterations
directly):

```{r salsoA}
best <- getBestChain(fit_k3)
n_saved <- nrow(best$samples)
post_labels <- best$samples[(n_saved / 2 + 1):n_saved, ]

point_est <- salso(post_labels, loss = "binder", maxNClusters = K_true)
table(point_est, true_labels)
```

The recovered partition matches the true one almost exactly.

## Part B: a case where the Gaussian assumption is insufficient

Generate data whose **true** generative process is heavy-tailed (a
multivariate $t$ with $\nu = 2.5$ degrees of freedom - few enough that
extreme values are common), then fit both an MVN and an MVT sampler to the
*same* data and compare.

```{r simB}
set.seed(505)

N <- 500
P <- 2
K <- 2
B <- 2
mu_true <- rbind(c(0, 0), c(6, 6))
batch_shift_true <- rbind(c(0, 0), c(0.4, 0.4))
df_true <- 2.5

labels <- sample(0:(K - 1), N, replace = TRUE)
batch_vec <- sample(0:(B - 1), N, replace = TRUE)

X <- matrix(0, N, P)
for (i in seq_len(N)) {
  k <- labels[i] + 1
  b <- batch_vec[i] + 1
  # Multivariate t via a chi-square scale mixture of a standard normal.
  w <- rchisq(1, df_true) / df_true
  X[i, ] <- mu_true[k, ] + batch_shift_true[b, ] + rnorm(P) / sqrt(w)
}
```

```{r simBviz}
ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], group = factor(labels)),
       aes(X1, X2, colour = group)) +
  geom_point(alpha = 0.6) +
  labs(title = "Heavy-tailed (t, df = 2.5) simulated data", subtitle = "Visible extreme values the MVN sampler will have to (mis)explain via inflated covariance") +
  theme_minimal()
```

```{r fitB}
n_iter <- 1500
thin <- 10
n_chains <- 4

out_mvn <- fitBatchMix(X, n_chains, n_iter, thin, batch_vec, "MVN", K_max = K)
out_mvt <- fitBatchMix(X, n_chains, n_iter, thin, batch_vec, "MVT", K_max = K)
```

```{r printB}
out_mvn
out_mvt
```

```{r processB}
proc_mvn <- processMCMCChain(getBestChain(out_mvn), burn = n_iter / 2)
proc_mvt <- processMCMCChain(getBestChain(out_mvt), burn = n_iter / 2)
```

```{r resultB_means}
cat("True cluster means:\n"); print(mu_true)
cat("MVN estimated means:\n"); print(t(proc_mvn$mean_est))
cat("MVT estimated means:\n"); print(t(proc_mvt$mean_est))
```

```{r resultB_cov}
cat("MVN estimated cluster covariance (diagonal):\n")
print(sapply(1:K, function(k) diag(proc_mvn$cov_est[, , k])))
cat("MVT estimated cluster covariance (diagonal):\n")
print(sapply(1:K, function(k) diag(proc_mvt$cov_est[, , k])))
cat("(true within-cluster variance of a t(df = 2.5) is df/(df-2) =", df_true / (df_true - 2), ")\n")
```

```{r resultB_labels}
acc <- function(pred, truth) {
  # accounts for the arbitrary label permutation at K = 2
  tab <- table(pred, truth)
  max(sum(diag(tab)), sum(tab) - sum(diag(tab))) / sum(tab)
}

cat("MVN label accuracy:", acc(proc_mvn$pred, labels), "\n")
cat("MVT label accuracy:", acc(proc_mvt$pred, labels), "\n")
```

**MVN's clustering essentially fails** (label accuracy near chance for
$K=2$) - its Gaussian likelihood tries to "explain" the heavy tails by
inflating the estimated covariance far past the true value, which destroys
the separation between clusters. MVT correctly attributes the extreme
values to heavy tails via its degrees-of-freedom parameter, recovers a
covariance close to the true value, and correctly labels the large
majority of items. This is the concrete case where the MVN assumption is
not just imprecise but actively misleading, and where MVT's extra
flexibility is worth its added cost.

## Part C: recovering a known correlation structure (MVN_LKJ)

Unlike `generateBatchData()`, which simulates independent columns by
design, recovering correlation needs data simulated directly. Cluster 1
has real correlation between features 1-2 and between 3-4; cluster 2 has
none at all, a check against false positives (finding correlation that
isn't there).

```{r simC}
set.seed(303)

N <- 500
P <- 4
K <- 2
B <- 2

R_true <- list(
  matrix(c(
    1, 0.7, 0, 0,
    0.7, 1, 0, 0,
    0, 0, 1, -0.5,
    0, 0, -0.5, 1
  ), 4, 4),
  diag(4)
)
sigma_true <- rbind(c(1, 1, 1, 1), c(1.2, 1.2, 1.2, 1.2))
mu_true <- rbind(c(0, 0, 0, 0), c(4, 4, 3, 3))
batch_shift_true <- rbind(c(0, 0, 0, 0), c(0.4, -0.3, 0.2, 0.1))

labels <- sample(0:(K - 1), N, replace = TRUE)
batch_vec <- sample(0:(B - 1), N, replace = TRUE)

X <- matrix(0, N, P)
for (i in seq_len(N)) {
  k <- labels[i] + 1
  b <- batch_vec[i] + 1
  Sigma_k <- diag(sigma_true[k, ]) %*% R_true[[k]] %*% diag(sigma_true[k, ])
  X[i, ] <- mu_true[k, ] + batch_shift_true[b, ] +
    as.numeric(MASS::mvrnorm(1, rep(0, P), Sigma_k))
}

ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], cluster = factor(labels)), aes(X1, X2, colour = cluster)) +
  geom_point(alpha = 0.7) +
  labs(title = "Simulated data (features 1-2)", subtitle = "Cluster 1 has real correlation between X1/X2; cluster 2 has none") +
  theme_minimal()
```

`type = "MVN_LKJ"` is fit the same way as `"MVN"`/`"MVT"`, through
`fitBatchMix()`, with auto-tuning left on its default - manually tuning
`r_proposal_window`/`sigma_proposal_window` (bundled into `control`, see
`?batchmixControl`, in place of `cov_proposal_window`) is not required for
typical use, including here:

```{r fitC}
fit_lkj <- fitBatchMix(
  X, n_chains = 4, n_iter = 4000, thin = 20, batch_vec, "MVN_LKJ",
  K_max = K, initial_labels = labels, fixed = NULL, eta = 1.0
)
fit_lkj
```

```{r resultsC}
chain <- getBestChain(fit_lkj)
n_saved <- dim(chain$covariance)[3]
last_half <- (n_saved / 2 + 1):n_saved

avg_R_by_k <- lapply(1:K, function(k) {
  Rs <- lapply(last_half, function(s) cov2cor(chain$covariance[, ((k - 1) * P + 1):(k * P), s]))
  Reduce(`+`, Rs) / length(Rs)
})

cat("Cluster 1 - posterior mean correlation:\n"); print(round(avg_R_by_k[[1]], 2))
cat("Cluster 1 - TRUE correlation:\n"); print(R_true[[1]])

cat("\nCluster 2 - posterior mean correlation:\n"); print(round(avg_R_by_k[[2]], 2))
cat("Cluster 2 - TRUE correlation (identity):\n"); print(R_true[[2]])
```

Cluster 1's strong correlations (0.7, -0.5) are recovered with the correct
sign and the correct relative ordering; cluster 2, which has **no** true
correlation, correctly shows no spurious structure (every off-diagonal
entry near 0) - exactly the check an Inverse-Wishart prior makes harder to
trust, since IW's coupling between scale and correlation means a
badly-set scale prior can leak into spurious apparent correlation.

### The effect of $\eta$

```{r eta_effect}
fit_lkj_eta5 <- fitBatchMix(
  X, n_chains = 4, n_iter = 4000, thin = 20, batch_vec, "MVN_LKJ",
  K_max = K, initial_labels = labels, fixed = NULL, eta = 5.0, verbose = FALSE
)

chain_eta5 <- getBestChain(fit_lkj_eta5)
avg_R_eta5_c1 <- Reduce(`+`, lapply(last_half, function(s) {
  cov2cor(chain_eta5$covariance[, 1:P, s])
})) / length(last_half)

cat("Cluster 1, eta = 1 (uniform prior) - estimated correlation:\n")
print(round(avg_R_by_k[[1]], 2))
cat("Cluster 1, eta = 5 (shrinks toward independence) - estimated correlation:\n")
print(round(avg_R_eta5_c1, 2))
```

Raising $\eta$ pulls the posterior correlation estimates towards zero, as
the LKJ theory predicts - useful as a weakly-informative regularizer when
$P$ is large relative to the amount of data per cluster and unconstrained
correlation estimates would be unstable.

## Summary

- `type = "MVN"` is the default, and given the true $K$ recovers a
  near-perfect partition when the data is (near-)Gaussian and the
  cluster structure is genuinely present (Part A). Comparing raw BIC
  across candidate $K$ values is a useful rough guide but not a decisive
  test on its own - cross-check against posterior cluster occupancy and
  domain knowledge.
- `type = "MVT"` matters specifically when the data has heavier tails than
  a Gaussian can represent - under such contamination MVN's clustering can
  fail outright, while MVT's degrees-of-freedom parameter absorbs the
  extreme values and recovers the correct structure (Part B).
- `type = "MVN_LKJ"` matters when the correlation *structure* itself is of
  interest, or when an Inverse-Wishart prior's coupling between scale and
  correlation is a concern; its LKJ concentration `eta` trades off between
  a uniform prior over correlation matrices (`eta = 1`) and shrinkage
  towards independence (`eta > 1`) (Part C).
- Every fit above was run as several chains via `fitBatchMix()` and
  inspected with `print()`/`getBestChain()` before reading off a
  conclusion - see the getting-started vignette for the fuller diagnostic
  and predictive-check bracket this abbreviates.
