---
title: "MVN vs MVT batch-corrected mixtures: model selection and robustness"
author: "batchmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVN vs MVT batch-corrected mixtures: model selection and robustness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Purpose

This vignette validates two things against data with a **known ground
truth**, and is grouped as one vignette because the two densities (MVN,
MVT) share the same batch-correction machinery and are directly comparable
fits to the same data:

1. That the (now-corrected) BIC criterion picks the right number of
   clusters $K$ when it is genuinely present in the data.
2. A concrete case where the Gaussian assumption is **insufficient**:
   heavy-tailed (outlier-contaminated) data, where the MVT sampler's
   robustness recovers the true structure and the MVN sampler does not.

Ground truth here comes from simulation, not an external "gold standard"
dataset. For validating whether a sampler recovers a *known* generative
model, this is in fact the strictest possible test — the alternative
(a real dataset with expert-assigned labels) has its own label noise, so a
sampler could look "wrong" for reasons that have nothing to do with the
implementation.

## The hierarchical model

For item $i = 1, \dots, N$, with cluster label $c_i \in \{1,\dots,K\}$ and
batch label $b_i \in \{1,\dots,B\}$ (both taken as **known** here, or with
$c_i$ inferred under a Dirichlet-multinomial prior when unsupervised):

$$
x_i \mid c_i = k,\ b_i = b \ \sim\
\begin{cases}
\mathcal{N}\!\left(\mu_k + m_b,\ S_b^{1/2} \Sigma_k S_b^{1/2}\right) & \text{(MVN)} \\[4pt]
t_{\nu_k}\!\left(\mu_k + m_b,\ S_b^{1/2} \Sigma_k S_b^{1/2}\right) & \text{(MVT)}
\end{cases}
$$

where $S_b = \mathrm{diag}(S_{b,1}, \dots, S_{b,P})$ is a per-batch,
per-feature multiplicative scale, $m_b$ is a per-batch additive shift, and
$\mu_k, \Sigma_k$ are the cluster mean and covariance. Priors:

$$
\Sigma_k \sim \mathrm{IW}(\nu, \Psi), \qquad
\mu_k \mid \Sigma_k \sim \mathcal{N}\!\left(\xi, \Sigma_k / \kappa\right), \qquad
m_b \sim \mathcal{N}(0, \delta^2 \lambda^2), \qquad
S_{b,p} - 1 \sim \mathrm{InvGamma}(\rho, \theta),
$$

$$
w \sim \mathrm{Dirichlet}(\alpha), \qquad
\nu_k - \nu_{\min} \sim \mathrm{Gamma}(\psi, \chi) \ \text{(MVT only)}.
$$

Model comparison across $K$ uses the (corrected) Bayesian information
criterion,

$$
\mathrm{BIC} = 2 \log \hat{L} - \left(p_{\text{cluster}} K_{\text{occ}} + p_{\text{batch}} B\right) \log N,
$$

with $p_{\text{cluster}} = 1 + P + P(P+1)/2$ (weight, mean, covariance)
and $p_{\text{batch}} = 2P$ (shift, scale) — see the package's development
notes for why this needed correcting from an earlier version that dropped
the cluster term entirely.

```{r libs}
library(batchmix)
library(salso)
library(ggplot2)
```

## Part A: does BIC recover the true K?

```{r simA}
set.seed(101)

N <- 600
P <- 2
K_true <- 3
B <- 4

group_means <- seq(1, K_true) * 4
std_dev <- rep(1.0, K_true)
batch_shift <- rnorm(B, mean = 0.3, sd = 0.2)
batch_var <- rep(1.3, B)

sim <- generateBatchData(
  N, P, group_means, std_dev, batch_shift, batch_var,
  rep(1 / K_true, K_true), rep(1 / B, B),
  type = "MVN"
)

X <- sim$observed_data
true_labels <- sim$group_IDs
batch_vec <- sim$batch_IDs
```

```{r simAviz}
ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], group = factor(true_labels), batch = factor(batch_vec)),
       aes(X1, X2, colour = group, shape = batch)) +
  geom_point(alpha = 0.7) +
  labs(title = "Simulated data: 3 true groups across 4 batches") +
  theme_minimal()
```

Fit at a range of candidate $K$ and compare BIC (higher is better in this
package's convention):

```{r bicA}
R_iter <- 1500
thin <- 5
n_chains <- 2

bic_by_k <- sapply(2:5, function(K) {
  out <- runMCMCChains(X, n_chains, R_iter, thin, batch_vec, "MVN", K = K)
  mean(sapply(out, function(o) {
    n <- length(o$BIC)
    mean(o$BIC[(n / 2):n])
  }))
})
names(bic_by_k) <- 2:5
round(bic_by_k, 1)
```

```{r bicA_result}
cat("K chosen by BIC:", names(which.max(bic_by_k)), " | true K:", K_true, "\n")
```

BIC correctly identifies $K = 3$. Note this depends on the clusters being
genuinely separable at the sample size given — with less separation (try
`group_means <- seq(1, K_true) * 2.5` above) BIC can conservatively prefer
fewer clusters when the extra cluster's likelihood gain does not clear its
complexity penalty; that is the criterion doing its job, not a fault.

Point-estimate the clustering at the true $K$ (via `salso`, which handles
the label-switching that any unsupervised mixture sampler exhibits — do
**not** compare raw per-iteration labels across chains or iterations
directly):

```{r salsoA}
out3 <- runMCMCChains(X, n_chains, R_iter, thin, batch_vec, "MVN", K = K_true)
samp <- out3[[1]]$samples
n_saved <- nrow(samp)
post_labels <- samp[(n_saved / 2 + 1):n_saved, ]

point_est <- salso(post_labels, loss = "binder", maxNClusters = K_true)
table(point_est, true_labels)
```

The recovered partition matches the true one almost exactly.

## Part B: a case where the Gaussian assumption is insufficient

Generate data whose **true** generative process is heavy-tailed (a
multivariate $t$ with $\nu = 2.5$ degrees of freedom — few enough that
extreme values are common), then fit both an MVN and an MVT sampler to the
*same* data and compare.

```{r simB}
set.seed(505)

N <- 500
P <- 2
K <- 2
B <- 2
mu_true <- rbind(c(0, 0), c(6, 6))
batch_shift_true <- rbind(c(0, 0), c(0.4, 0.4))
df_true <- 2.5

labels <- sample(0:(K - 1), N, replace = TRUE)
batch_vec <- sample(0:(B - 1), N, replace = TRUE)

X <- matrix(0, N, P)
for (i in seq_len(N)) {
  k <- labels[i] + 1
  b <- batch_vec[i] + 1
  # Multivariate t via a chi-square scale mixture of a standard normal.
  w <- rchisq(1, df_true) / df_true
  X[i, ] <- mu_true[k, ] + batch_shift_true[b, ] + rnorm(P) / sqrt(w)
}
```

```{r simBviz}
ggplot(data.frame(X1 = X[, 1], X2 = X[, 2], group = factor(labels)),
       aes(X1, X2, colour = group)) +
  geom_point(alpha = 0.6) +
  labs(title = "Heavy-tailed (t, df = 2.5) simulated data", subtitle = "Visible extreme values the MVN sampler will have to (mis)explain via inflated covariance") +
  theme_minimal()
```

```{r fitB}
R_iter <- 1500
thin <- 5
n_chains <- 2

out_mvn <- runMCMCChains(X, n_chains, R_iter, thin, batch_vec + 1, "MVN", K = K)
out_mvt <- runMCMCChains(X, n_chains, R_iter, thin, batch_vec + 1, "MVT", K = K)

proc_mvn <- processMCMCChains(out_mvn, burn = R_iter / 2)[[1]]
proc_mvt <- processMCMCChains(out_mvt, burn = R_iter / 2)[[1]]
```

```{r resultB_means}
cat("True cluster means:\n"); print(mu_true)
cat("MVN estimated means:\n"); print(t(proc_mvn$mean_est))
cat("MVT estimated means:\n"); print(t(proc_mvt$mean_est))
```

```{r resultB_cov}
cat("MVN estimated cluster covariance (diagonal):\n")
print(sapply(1:K, function(k) diag(proc_mvn$cov_est[, , k])))
cat("MVT estimated cluster covariance (diagonal):\n")
print(sapply(1:K, function(k) diag(proc_mvt$cov_est[, , k])))
cat("(true within-cluster variance of a t(df = 2.5) is df/(df-2) =", df_true / (df_true - 2), ")\n")
```

```{r resultB_labels}
acc <- function(pred, truth) {
  # accounts for the arbitrary label permutation at K = 2
  tab <- table(pred, truth)
  max(sum(diag(tab)), sum(tab) - sum(diag(tab))) / sum(tab)
}

samp_mvn <- out_mvn[[1]]$samples; n_saved <- nrow(samp_mvn)
pt_mvn <- salso(samp_mvn[(n_saved / 2 + 1):n_saved, ], loss = "binder", maxNClusters = K)

samp_mvt <- out_mvt[[1]]$samples
pt_mvt <- salso(samp_mvt[(n_saved / 2 + 1):n_saved, ], loss = "binder", maxNClusters = K)

cat("MVN label accuracy:", acc(pt_mvn, labels), "\n")
cat("MVT label accuracy:", acc(pt_mvt, labels), "\n")
```

**MVN's clustering essentially fails** (label accuracy near chance for
$K=2$) — its Gaussian likelihood tries to "explain" the heavy tails by
inflating the estimated covariance far past the true value, which
destroys the separation between clusters. MVT correctly attributes the
extreme values to heavy tails via its degrees-of-freedom parameter,
recovers a covariance close to the true value, and correctly labels the
large majority of items. This is the concrete case where the MVN
assumption is not just imprecise but actively misleading, and where MVT's
extra flexibility is worth its added cost.
