---
title: "Mixed continuous/binary data with missing and censored observations"
author: "batchmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed continuous/binary data with missing and censored observations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Purpose

This is a genuinely different model from MVN/MVT, not a variant of them:
it lets some columns be **binary** (e.g. a seropositive/negative call)
observed through a probit link, and lets any continuous column contain
**missing** or **censored** entries (e.g. below an assay's detection
limit) — none of which a plain multivariate-normal likelihood can
represent at all, no matter how it is tuned. That is the concrete sense
in which MVN/MVT are insufficient here: this is a data-type limitation,
not a robustness one.

## The hierarchical model

Every column, continuous or binary, is generated from **one shared**
latent Gaussian vector (Dunson, 2000, *JRSS-B*; Dunson & Herring, 2005,
*Biostatistics*):

$$
z_i \mid c_i = k,\ b_i = b \ \sim\ \mathcal{N}\!\left(\mu_k + m_b,\ \Sigma_k^{(b)}\right),
\qquad \Sigma_k^{(b)} = \Delta_b \left(\Delta_k R_k \Delta_k\right) \Delta_b,
$$

and each column $p$ has its own observation model applied to $z_{i,p}$:

$$
y_{i,p} =
\begin{cases}
z_{i,p} & p \text{ continuous, observed} \\
\text{unobserved} & p \text{ missing} \\
\text{known bound } c,\text{ with } z_{i,p} < c \text{ or } z_{i,p} > c & p \text{ censored} \\
\mathbb{1}\{z_{i,p} > 0\} & p \text{ binary (probit link)}
\end{cases}
$$

**Identifiability** (Chib & Greenberg, 1998, *Biometrika*): a probit
model is identified only up to the *correlation* matrix of $z$ — rescaling
one coordinate's variance leaves $\mathrm{sign}(z)$, and hence the
observed binary outcome, unchanged. So for a binary column $p$, both scale
layers are fixed: $\sigma_{k,p} = 1$ for every cluster $k$ and
$S_{b,p} = 1$ for every batch $b$; only $R_k$ (correlation, including
binary-binary and binary-continuous entries — these are the classical
tetrachoric and polyserial correlations, both identified) and the
location parameters $\mu_{k,p}, m_{b,p}$ are estimated for that column.

Inference proceeds by Gibbs data augmentation (Albert & Chib, 1993,
*JASA*): given the current parameters, each missing/censored/binary entry
of $z_i$ is redrawn from its Normal conditional on every other current
entry of the same item (via the precision matrix, already computed each
sweep) — a plain draw if missing, truncated to the implied half-line if
binary, truncated against the recorded bound if censored. Every other
parameter update then proceeds exactly as in the fully-continuous
sampler, operating on the completed $z$.

```{r libs}
library(batchmix)
library(ggplot2)
```

## Part A: correlation recovery with mixed continuous/binary columns

To isolate the specific new capability being tested, this uses a single
cluster ($K=1$) — clustering with label-switching is validated separately
in the other vignettes, and combining every new feature into one maximally
complex scenario would make a poor fit hard to distinguish from a genuine
bug. Note also that vanilla Gibbs sampling for multivariate probit
correlations is known to mix slowly (the augmented data and the
correlation matrix are highly correlated in the posterior) — this shows up
below as needing a smaller Metropolis step (`r_proposal_window`) and a
longer chain than the purely continuous LKJ vignette needed. Parameter
expansion (Imai & van Dyk, 2005, *Political Analysis*; Talhouk, Doucet &
Murphy, 2012, *JCGS*) is the standard fix if this matters in practice; it
is not implemented here.

```{r simA}
set.seed(7)

N <- 400
P <- 4 # columns 1-2 continuous, 3-4 binary
R_true <- matrix(c(
  1, 0.6, 0, 0,
  0.6, 1, 0, 0,
  0, 0, 1, 0.7,
  0, 0, 0.7, 1
), 4, 4)

L <- chol(R_true)
Z <- matrix(rnorm(N * P), N, P) %*% L
X <- Z
X[, 3:4] <- (Z[, 3:4] > 0) * 1

column_type <- c(0L, 0L, 1L, 1L)
censor_code <- matrix(0L, N, P)
labels0 <- rep(0L, N)
batch_vec0 <- rep(0L, N)
```

```{r simVizA}
viz_df <- data.frame(X1 = X[, 1], X2 = X[, 2], binary3 = factor(X[, 3]), binary4 = factor(X[, 4]))

ggplot(viz_df, aes(X1, X2, colour = binary4)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Continuous columns 1-2, coloured by binary column 4",
    subtitle = "Column 4 is correlated with column 3 (not shown), not with 1-2 - the model must recover this from the data alone"
  ) +
  theme_minimal()
```

```{r fitA}
out <- sampleMVNMixed(
  X, 1, 1, labels0, batch_vec0, column_type, censor_code,
  mu_proposal_window = 0.3, r_proposal_window = 0.03, sigma_proposal_window = 40,
  m_proposal_window = 0.2, S_proposal_window = 40,
  R = 8000, thin = 20, concentration = 1,
  m_scale = 0.01, rho = 3, theta = 1, eta = 1.0, sample_m_scale = FALSE,
  auto_tune = FALSE, n_burn = 0,
  include_interaction = FALSE, gamma_proposal_window = 0.1, a_gamma = 2.0, b_gamma = 1.0,
  weight_prior_type = 0, batch_coordinates = numeric(0),
  gp_tau2 = 1.0, gp_length_scale = 1.0, eta_proposal_window = 0.1,
  sample_gp_hyperparameters = FALSE, gp_hyperparameter_proposal_window = 0.1,
  pp_tau2_shape = 2.0, pp_tau2_rate = 1.0, pp_mu_prior_sd = 10.0
)

n_saved <- dim(out$covariance)[3]
last_half <- (n_saved / 2 + 1):n_saved
r12 <- sapply(last_half, function(s) cov2cor(out$covariance[, 1:P, s])[1, 2])
r34 <- sapply(last_half, function(s) cov2cor(out$covariance[, 1:P, s])[3, 4])
```

```{r resultA}
cat("r_acceptance_rate:", out$r_acceptance_rate, "\n")
cat("Continuous-continuous correlation (cols 1-2), true = 0.6:\n")
cat("  posterior mean:", mean(r12), " range over kept samples:", range(r12), "\n")
cat("Binary-binary correlation (cols 3-4), true = 0.7:\n")
cat("  posterior mean:", mean(r34), " range over kept samples:", range(r34), "\n")
```

Both correlations are recovered with the correct sign and roughly the
right magnitude. The binary-binary estimate is visibly noisier across the
kept samples than the continuous-continuous one — exactly the slow-mixing
symptom flagged above, not a separate bug; a longer chain or a smaller
`r_proposal_window` tightens it further (verified separately, not
reproduced here to keep this vignette's runtime reasonable).

## Part B: does proper missing/censoring handling matter?

Compare the sampler's data augmentation against the common naive
approach — taking a censored value at its recorded (bound) value, exactly
as it arrives in the raw data — on a case where the censoring is common
enough that the naive approach's bias should be visible.

```{r simB}
set.seed(606)

N <- 400
mu_true <- c(0, 3)
Xb <- matrix(rnorm(N * 2), N, 2)
Xb[, 1] <- Xb[, 1] + mu_true[1]
Xb[, 2] <- Xb[, 2] + mu_true[2]

# Left-censor column 2 at its own mean, so roughly half the data is
# censored - a deliberately severe case to make the bias clearly visible.
censor_code_b <- matrix(0L, N, 2)
left_cens <- which(Xb[, 2] < 3.0)
cat("Fraction of column 2 censored:", length(left_cens) / N, "\n")
censor_code_b[left_cens, 2] <- 1L
Xb[left_cens, 2] <- 3.0

column_type_b <- c(0L, 0L)
```

```{r fitB}
out_b <- sampleMVNMixed(
  Xb, 1, 1, labels0, batch_vec0, column_type_b, censor_code_b,
  mu_proposal_window = 0.2, r_proposal_window = 0.1, sigma_proposal_window = 40,
  m_proposal_window = 0.2, S_proposal_window = 40,
  R = 8000, thin = 10, concentration = 1,
  m_scale = 0.01, rho = 3, theta = 1, eta = 1.0, sample_m_scale = FALSE,
  auto_tune = FALSE, n_burn = 0,
  include_interaction = FALSE, gamma_proposal_window = 0.1, a_gamma = 2.0, b_gamma = 1.0,
  weight_prior_type = 0, batch_coordinates = numeric(0),
  gp_tau2 = 1.0, gp_length_scale = 1.0, eta_proposal_window = 0.1,
  sample_gp_hyperparameters = FALSE, gp_hyperparameter_proposal_window = 0.1,
  pp_tau2_shape = 2.0, pp_tau2_rate = 1.0, pp_mu_prior_sd = 10.0
)

n_saved_b <- dim(out_b$means)[3]
# A generous burn-in: the chain's initial state is a prior draw, which can
# start a considerable distance from the posterior mode (see the trace
# below) - the mean location prior here is deliberately diffuse (kappa is
# small), so this is expected, not a sign of a problem.
keep_b <- (0.75 * n_saved_b + 1):n_saved_b
mu_proper <- rowMeans(out_b$means[, 1, keep_b])
mu_naive <- colMeans(Xb)
```

```{r resultB}
cat("True mu:", mu_true, "\n")
cat("Proper (augmented) estimate:", round(mu_proper, 3), "\n")
cat("Naive (censored values taken at face value) estimate:", round(mu_naive, 3), "\n")
```

```{r resultB_trace, fig.width=6, fig.height=3}
plot(out_b$means[2, 1, ], type = "l", xlab = "saved iteration", ylab = "mu (column 2)",
     main = "Convergence to the true value (dashed) from a diffuse prior start")
abline(h = mu_true[2], lty = 2, col = "red")
abline(v = 0.75 * n_saved_b, lty = 3, col = "grey40")
```

The naive estimate is biased upward by the censoring (values below the
bound get replaced by a value that is, on average, too high); the proper
augmentation recovers a value much closer to the truth. The trace plot
also makes explicit why an adequate burn-in matters here — this is
ordinary MCMC practice, not something specific to this sampler, but it is
easy to get wrong by eye when a chain's early transient looks superficially
like noise rather than a systematic drift towards the posterior mode.
