---
title: "Mixed continuous/binary data with missing and censored observations"
author: "batchmix"
output: rmarkdown::html_vignette
description: |
  Fitting type = "MVN_MIXED": data with binary/probit columns and censored
  continuous entries, neither of which MVN/MVT/MVN_LKJ can represent -
  checked against a known correlation structure and against the bias a
  naive treatment of censored data would introduce.
vignette: >
  %\VignetteIndexEntry{Mixed continuous/binary data with missing and censored observations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
```

## Purpose

`type = "MVN_MIXED"` is a genuinely different model from MVN/MVT/MVN_LKJ,
not a variant of them: it lets some columns be **binary** (e.g. a
seropositive/negative call) observed through a probit link, and lets any
continuous column contain **censored** entries (e.g. below an assay's
detection limit) - neither of which a plain multivariate-normal likelihood
can represent at all, no matter how it is tuned. That is the concrete
sense in which the other three types are insufficient here: this is a
data-type limitation, not a robustness one.

Plain **missing** continuous entries (no binary columns, no censoring) are
*not* one of those data-type limitations: `MVN`, `MVT` and `MVN_LKJ` each
model missing entries of `X` directly, via the same per-sweep Gibbs data
augmentation this vignette's `MVN_MIXED` examples use - so for purely
continuous data with some missing cells, prefer whichever of those three
densities otherwise fits the data (see the covariance-model vignette) over
`MVN_MIXED`, and reach for `MVN_MIXED` specifically when the data
additionally has binary/probit columns or genuinely censored (not merely
missing) entries. Prior/posterior predictive checks (see the
getting-started vignette) are supported for `type = "MVN_MIXED"` too
(`simulatePriorPredictive()`/`simulatePosteriorPredictive()`,
`column_type` argument): a binary column's replicate is the same shared
latent Gaussian draw thresholded at 0 (Albert & Chib, 1993), matching
`sigma`/`S` being fixed at 1 for that column below. Every replicate is
fully observed - a censored cell's recorded value is only a bound, not the
true value, so replicating "what a censored draw would look like" would
mean also modelling the censoring mechanism itself, which is not attempted
here; `plotPredictiveCheck()`'s `censor_code` argument instead excludes
those cells from the observed side of the comparison. Used alongside the
MCMC convergence diagnostics in this vignette, not in place of them.

## The hierarchical model

Every column, continuous or binary, is generated from **one shared**
latent Gaussian vector (Dunson, 2000, *JRSS-B*; Dunson & Herring, 2005,
*Biostatistics*):

$$
z_i \mid c_i = k,\ b_i = b \ \sim\ \mathcal{N}\!\left(\mu_k + m_b,\ \Sigma_k^{(b)}\right),
\qquad \Sigma_k^{(b)} = \Delta_b \left(\Delta_k R_k \Delta_k\right) \Delta_b,
$$

and each column $p$ has its own observation model applied to $z_{i,p}$:

$$
y_{i,p} =
\begin{cases}
z_{i,p} & p \text{ continuous, observed} \\
\text{unobserved} & p \text{ missing} \\
\text{known bound } c,\text{ with } z_{i,p} < c \text{ or } z_{i,p} > c & p \text{ censored} \\
\mathbb{1}\{z_{i,p} > 0\} & p \text{ binary (probit link)}
\end{cases}
$$

**Identifiability** (Chib & Greenberg, 1998, *Biometrika*): a probit model
is identified only up to the *correlation* matrix of $z$ - rescaling one
coordinate's variance leaves $\mathrm{sign}(z)$, and hence the observed
binary outcome, unchanged. So for a binary column $p$, both scale layers
are fixed: $\sigma_{k,p} = 1$ for every cluster $k$ and $S_{b,p} = 1$ for
every batch $b$; only $R_k$ (correlation, including binary-binary and
binary-continuous entries - these are the classical tetrachoric and
polyserial correlations, both identified) and the location parameters
$\mu_{k,p}, m_{b,p}$ are estimated for that column.

Inference proceeds by Gibbs data augmentation (Albert & Chib, 1993,
*JASA*): given the current parameters, each missing/censored/binary entry
of $z_i$ is redrawn from its Normal conditional on every other current
entry of the same item (via the precision matrix, already computed each
sweep) - a plain draw if missing, truncated to the implied half-line if
binary, truncated against the recorded bound if censored. Every other
parameter update then proceeds exactly as in the fully-continuous
sampler, operating on the completed $z$.

```{r libs, message=FALSE}
library(batchmix)
library(ggplot2)
```

## Part A: correlation recovery with mixed continuous/binary columns

To isolate the specific new capability being tested, this uses a single
cluster and a single batch - clustering with label-switching is validated
separately in the other vignettes, and combining every new feature into
one maximally complex scenario would make a poor fit hard to distinguish
from a genuine bug. Vanilla Gibbs sampling for multivariate probit
correlations is also known to mix more slowly than the purely continuous
samplers (the augmented data and the correlation matrix are highly
correlated in the posterior), so this fit uses a smaller, hand-set
`r_proposal_window` (bundled into `control`, with auto-tuning switched
off) rather than the defaults the other vignettes leave untouched -
`?batchmixControl` covers this as the situation manual tuning is for. Parameter expansion (Imai & van Dyk, 2005, *Political Analysis*;
Talhouk, Doucet & Murphy, 2012, *JCGS*) is the standard fix if slow mixing
matters in practice; it is not implemented here.

```{r simA}
set.seed(7)

N <- 400
P <- 4 # columns 1-2 continuous, 3-4 binary
R_true <- matrix(c(
  1, 0.6, 0, 0,
  0.6, 1, 0, 0,
  0, 0, 1, 0.7,
  0, 0, 0.7, 1
), 4, 4)

L <- chol(R_true)
Z <- matrix(rnorm(N * P), N, P) %*% L
X <- Z
X[, 3:4] <- (Z[, 3:4] > 0) * 1

column_type <- c(0L, 0L, 1L, 1L)
censor_code <- matrix(0L, N, P)
batch_vec0 <- rep(0L, N)
```

```{r simVizA}
viz_df <- data.frame(X1 = X[, 1], X2 = X[, 2], binary3 = factor(X[, 3]), binary4 = factor(X[, 4]))

ggplot(viz_df, aes(X1, X2, colour = binary4)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Continuous columns 1-2, coloured by binary column 4",
    subtitle = "Column 4 is correlated with column 3 (not shown), not with 1-2 - the model must recover this from the data alone"
  ) +
  theme_minimal()
```

```{r fitA}
fit_a <- fitBatchMix(
  X, n_chains = 4, n_iter = 8000, thin = 20, batch_vec0, "MVN_MIXED",
  K_max = 1, fixed = NULL, m_scale = 0.01, initial_labels = rep(0L, N),
  column_type = column_type, censor_code = censor_code,
  control = batchmixControl(
    mu_proposal_window = 0.3, r_proposal_window = 0.03, sigma_proposal_window = 40,
    m_proposal_window = 0.2, S_proposal_window = 40, auto_tune = FALSE
  )
)
fit_a
```

`m_scale` is fixed small here (rather than left to its default of being
estimated) because with a single batch there is no batch replication to
inform it - a small fixed value keeps the batch shift near zero so the
cluster mean, not an arbitrary mean/shift split, carries the location
signal. `getBestChain()` (by post burn-in mean BIC) is used to select
among the four chains below, not only as routine practice: with this
proposal window some chains do settle on a poorer fit than others, which
the BIC comparison correctly identifies.

```{r resultA}
chain_a <- getBestChain(fit_a)
n_saved <- dim(chain_a$covariance)[3]
last_half <- (n_saved / 2 + 1):n_saved
r12 <- sapply(last_half, function(s) cov2cor(chain_a$covariance[, 1:P, s])[1, 2])
r34 <- sapply(last_half, function(s) cov2cor(chain_a$covariance[, 1:P, s])[3, 4])
r12_mean <- mean(r12)
r34_mean <- mean(r34)

cat("Continuous-continuous correlation (cols 1-2), true = 0.6:\n")
cat("  posterior mean:", r12_mean, " range over kept samples:", range(r12), "\n")
cat("Binary-binary correlation (cols 3-4), true = 0.7:\n")
cat("  posterior mean:", r34_mean, " range over kept samples:", range(r34), "\n")

rhat_a <- attr(fit_a, "convergence")$rhat
```

```{r resultA_direction, echo=FALSE}
describe_recovery <- function(est, truth) {
  err <- est - truth
  if (abs(err) < 0.03) {
    "recovered close to its true value"
  } else if (err > 0) {
    paste0("overestimated by about ", round(err, 2))
  } else {
    paste0("underestimated by about ", round(-err, 2))
  }
}
r12_desc <- describe_recovery(r12_mean, 0.6)
r34_desc <- describe_recovery(r34_mean, 0.7)
```

Both correlations recover the correct **sign**; the continuous-continuous
correlation (cols 1-2) is `r r12_desc` (posterior mean
`r round(r12_mean, 2)` against a true value of 0.6), and the binary-binary
correlation (cols 3-4) is `r r34_desc` (posterior mean
`r round(r34_mean, 2)` against a true value of 0.7). R&#770; on this fit's
complete-data log-likelihood is `r round(rhat_a, 2)` - well above the
~1.01 rule of thumb, consistent with the slow probit-correlation mixing
flagged above, and a reminder that either correlation here is best read as
indicative of sign and approximate magnitude rather than as a precise
value: which of the two correlations is closer to its true value on any
given run is a property of exactly *how* the chains have not yet
converged, not a reliable property of the sampler that should be expected
to replicate on different data. Readers relying on `type = "MVN_MIXED"`
for a specific correlation estimate are encouraged to check convergence
(as here, and as the R&#770; above illustrates directly) rather than take a
single chain's answer at face value.

### Predictive checks

A binary column's replicate is the same shared latent Gaussian draw
thresholded at 0 - `style = "statistic"` with `statistic = mean` is the
right check for one (the proportion of 1s), not `style = "density"` (which
assumes a continuous column; `plotPredictiveCheck()` warns if asked to use
it on an apparently-binary column anyway).

```{r predCheckA, fig.width=6, fig.height=3.5}
prior_sims_a <- simulatePriorPredictive(
  X, batch_vec0, K = 1, type = "MVN_MIXED", column_type = column_type
)
plotPredictiveCheck(X, prior_sims_a,
  style = "density", column = 1
) # continuous column: density overlay is appropriate
plotPredictiveCheck(X, prior_sims_a,
  style = "statistic", column = 3,
  statistic = mean, statistic_name = "P(column 3 = 1)"
) # binary column: proportion, not density

post_sims_a <- simulatePosteriorPredictive(
  chain_a, batch_vec0,
  column_type = column_type, burn = 4000, n_draws = 100
)
plotPredictiveCheck(X, post_sims_a,
  style = "statistic", column = 4,
  statistic = mean, statistic_name = "P(column 4 = 1)"
)
```

The posterior predictive check uses `chain_a`'s own stored `mean_sum`/
`cov_comb` at each retained iteration directly - no separate prior/
empirical-Bayes formula is re-derived, so there is no risk of it drifting
out of sync with what the sampler actually fit.

## Part B: does proper missing/censoring handling matter?

Compare the sampler's data augmentation against the common naive
approach - taking a censored value at its recorded (bound) value, exactly
as it arrives in the raw data - on a case where the censoring is common
enough that the naive approach's bias should be visible.

```{r simB}
set.seed(606)

N <- 400
mu_true <- c(0, 3)
Xb <- matrix(rnorm(N * 2), N, 2)
Xb[, 1] <- Xb[, 1] + mu_true[1]
Xb[, 2] <- Xb[, 2] + mu_true[2]

# Left-censor column 2 at its own mean, so roughly half the data is
# censored - a deliberately severe case to make the bias clearly visible.
censor_code_b <- matrix(0L, N, 2)
left_cens <- which(Xb[, 2] < 3.0)
cat("Fraction of column 2 censored:", length(left_cens) / N, "\n")
censor_code_b[left_cens, 2] <- 1L
Xb[left_cens, 2] <- 3.0

column_type_b <- c(0L, 0L)
```

```{r fitB}
fit_b <- fitBatchMix(
  Xb, n_chains = 4, n_iter = 8000, thin = 10, batch_vec0, "MVN_MIXED",
  K_max = 1, fixed = NULL, m_scale = 0.01, initial_labels = rep(0L, N),
  column_type = column_type_b, censor_code = censor_code_b,
  control = batchmixControl(
    mu_proposal_window = 0.2, r_proposal_window = 0.1, sigma_proposal_window = 40,
    m_proposal_window = 0.2, S_proposal_window = 40, auto_tune = FALSE
  )
)
fit_b
```

As in Part A, some of the four chains settle on a poorer mean/shift split
than others; `getBestChain()` selects the one with the best post burn-in
BIC.

```{r resultB}
chain_b <- getBestChain(fit_b)
n_saved_b <- dim(chain_b$means)[3]
# A generous burn-in: the chain's initial state is a prior draw, which can
# start a considerable distance from the posterior mode (see the trace
# below) - the mean location prior here is deliberately diffuse (kappa is
# small), so this is expected, not a sign of a problem.
keep_b <- (0.75 * n_saved_b + 1):n_saved_b
mu_proper <- rowMeans(chain_b$means[, 1, keep_b])
mu_naive <- colMeans(Xb)

cat("True mu:", mu_true, "\n")
cat("Proper (augmented) estimate:", round(mu_proper, 3), "\n")
cat("Naive (censored values taken at face value) estimate:", round(mu_naive, 3), "\n")
```

```{r resultB_trace, fig.width=6, fig.height=3}
plot(chain_b$means[2, 1, ], type = "l", xlab = "saved iteration", ylab = "mu (column 2)",
     main = "Convergence to the true value (dashed) from a diffuse prior start")
abline(h = mu_true[2], lty = 2, col = "red")
abline(v = 0.75 * n_saved_b, lty = 3, col = "grey40")
```

```{r resultB_rhat, echo=FALSE}
rhat_b <- attr(fit_b, "convergence")$rhat
naive_err <- abs(mu_naive[2] - mu_true[2])
proper_err <- abs(mu_proper[2] - mu_true[2])
```

The naive estimate is biased upward by the censoring (values below the
bound get replaced by a value that is, on average, too high): it is off
by `r round(naive_err, 2)`, versus `r round(proper_err, 2)` for the
proper augmented estimate - a large reduction in error, even though this
fit's own R&#770; (`r round(rhat_b, 2)`) is, as in Part A, still well above
the ~1.01 rule of thumb. That combination is worth sitting with rather
than glossing over: a fit can be simultaneously (i) clearly, verifiably
better than the naive alternative on the one comparison this section is
about, and (ii) not yet fully converged by the package's own diagnostic.
Both are true here; neither contradicts the other, and it is why this
package reports R&#770; explicitly rather than leaving a reader to infer
"the point estimate looked sensible, so the chain must have converged."
The trace plot also makes explicit why an adequate burn-in matters here -
this is ordinary MCMC practice, not something specific to this sampler,
but it is easy to get wrong by eye when a chain's early transient looks
superficially like noise rather than a systematic drift towards the
posterior mode.

## Summary

- `type = "MVN_MIXED"` is for a data-type limitation (binary/probit
  columns, genuinely censored entries), not a robustness one - reach for
  it specifically when the data has one of those, not merely missing
  continuous cells (every type handles those).
- Both new capabilities were checked against a known ground truth: the
  correlation structure (Part A) and the bias a naive treatment of
  censored data would introduce (Part B).
- The probit correlation update mixes more slowly than the purely
  continuous samplers - this is the case in this package where manually
  tuning `control`'s proposal windows (rather than leaving `auto_tune =
  TRUE`) is worth doing, unlike the other vignettes.
- `simulatePriorPredictive()`/`simulatePosteriorPredictive()` support this
  `type` too (via `column_type`), alongside `print()`/`summary()` on the
  fit and standard MCMC trace/acceptance-rate diagnostics, for judging
  whether a fit of this type can be trusted.
